Optimal transport for automatic alignment of untargeted metabolomic data

Version 1

Abstract

Abstract

Untargeted metabolomic profiling through liquid chromatography-mass spectrometry (LC-MS) measures a vast array of metabolites within biospecimens, advancing drug development, disease diagnosis, and risk prediction. However, the low throughput of LC-MS poses a major challenge for biomarker discovery, annotation, and experimental comparison, necessitating the merging of multiple datasets. Current data pooling methods encounter practical limitations due to their vulnerability to data variations and hyperparameter dependence. Here we introduce GromovMatcher, a flexible and user-friendly algorithm that automatically combines LC-MS datasets using optimal transport. By capitalizing on feature intensity correlation structures, GromovMatcher delivers superior alignment accuracy and robustness compared to existing approaches. This algorithm scales to thousands of features requiring minimal hyperparameter tuning. Applying our method to experimental patient studies of liver and pancreatic cancer, we discover shared metabolic features related to patient alcohol intake, demonstrating how GromovMatcher facilitates the search for biomarkers associated with lifestyle risk factors linked to several cancer types.

Introduction

Untargeted metabolomics is a powerful analytical technique used to identify and measure a large number of metabolites in a biological sample without preselecting targets Patti (2011). This approach allows for a comprehensive overview of an individual’s metabolic profile, provides insights into the biochemical processes involved in cellular and organismal physiology Wishart (2019); Pirhaji et al. (2016), and allows for the exploration of how environmental factors impact metabolism Rappaport et al. (2014); Bedia (2022). It creates new opportunities to investigate health-related conditions, including diabetes Wang et al. (2011), inflammatory bowel diseases Franzosa et al. (2019), and various cancer types Loftfield et al. (2021); Li et al. (2020). However, a major challenge in biomarker discovery, metabolic signature identification and other untargeted metabolomic analyses lies in the low throughput of experimental data, necessitating the development of efficient pooling algorithms capable of merging datasets from multiple sources Loftfield et al. (2021).

A common experimental technique in untargeted metabolomics is liquid chromatography-mass spectrometry (LC-MS) which assembles a list of thousands of unlabeled metabolic features characterized by their mass-to-charge ratio (m/z), retention time (RT) Zhou et al. (2012), and intensity across all biological samples. Combining LC-MS datasets from multiple experimental studies remains challenging due to variation in the m/z and RT of a feature from one study to another Zhou et al. (2012); Ivanisevic and Want (2019). This problem is further compounded by differing instruments and analytical protocols across laboratories, resulting in seemingly incompatible metabolomic datasets.

Manual matching of metabolic features can be a laborious and error-prone task Loftfield et al. (2021). To address this challenge, several automated methods have been developed for metabolic feature alignment. One such method is MetaXCMS, which matches LC-MS features based on user-defined m/z and RT thresholds Tautenhahn et al. (2011). More advanced tools use information on feature intensities measured in samples. For instance, PAIRUP-MS uses known shared metabolic features to impute the intensities of all features from one dataset to another Hsu et al. (2019). MetabCombiner Habra et al. (2021) and M2S Climaco Pinto et al. (2022) compare average feature intensities, along with their m/z and RT values, to align datasets without requiring extensive knowledge of shared features. These automated alignment methods have accelerated our ability to pool and annotate datasets as well as extract biologically meaningful biomarkers. However, they demand substantial fine-tuning of user-defined parameters and ignore correlations among metabolic features which provide a wealth of additional information on shared features.

Here we introduce GromovMatcher, a user-friendly flexible algorithm which automates the matching of metabolic features across experiments. The main technical innovation of Gromov-Matcher lies in its ability to incorporate the correlation information between metabolic feature intensities, building upon the powerful mathematical framework of computational optimal transport (OT) Peyré et al. (2019); Villani (2021). OT has proven effective in solving various matching problems and has found applications in multiomics analysis Demetci et al. (2022), cell development Schiebinger et al. (2019); Yang et al. (2020), and chromatogram alignment Skoraczynski et al. (2022). Here we leverage the Gromov-Wasserstein (GW) method Memoli (2011); Solomon et al. (2016), which matches datasets based on their distance structure and has been seminally applied to spatial reconstruction problems in genomics Nitzan et al. (2019). GromovMatcher builds upon the GW algorithm to automatically uncover the shared correlation structure among metabolic feature intensities while also incorporating m/z and RT information in the final matching process.

To assess the performance of GromovMatcher, we systematically benchmark it on synthetic data with varying levels of noise, feature overlap, and data normalizations, outperforming prior state-of-the-art methods of metabCombiner Habra et al. (2021) and M2S Climaco Pinto et al. (2022). Next we apply GromovMatcher to align experimental patient studies of liver and pancreatic cancer to a reference dataset and associate the shared metabolic features to each patient’s alcohol intake. Through these efforts, we demonstrate how GromovMatcher data pooling improves our ability to discover biomarkers of lifestyle risk factors associated with several types of cancer.

Results

GromovMatcher algorithm

GromovMatcher uses the mathematical framework of OT to find all matching metabolic features between two untargeted metabolomic datasets (Fig. 1). It accepts two LC-MS datasets with possibly different numbers of metabolic features and samples. Each feature, fx، in Dataset 1 and fyj in Dataset 2, is identified by its m/z, RT, and vector of feature intensities across samples (Fig. 1a). The primary tenet of GromovMatcher is that shared metabolic features have similar correlation patterns in both datasets and can be matched based on the distance/correlations between their feature intensity vectors. Specifically, GromovMatcher computes the pairwise distances between the feature intensity vectors of each metabolic feature in a dataset and saves them into a distance matrix, one per dataset (Fig. 1b). In practice, we use either the Euclidean distance or the cosine distance (negative of correlation) to perform this step (Methods). The resulting distance matrices contain information about the feature intensity similarity within each study. Using optimal transport, we can deduce shared subsets of metabolic features in both datasets which have corresponding feature intensity distance structures.

An optimal transport approach for combining untargeted metabolomics datasets (GromovMatcher). (A) Inputs are two LC-MS datasets of unlabeled metabolic features (rows) identified by their m/z, RT, and feature intensities across biospecimen samples. Both studies can have differing numbers of metabolic features and samples. (B) In both datasets, the intensities across samples of each metabolic feature are formed into a vector and Euclidean distances between these feature vectors are computed and stored in a distance matrix. (C) Based on the technique of optimal transport, the unbalanced GW algorithm learns a coupling matrix ∏˜ that places large weights ∏ij˜≥0 when fxi and fyj likely correspond to the same metabolic feature. It optimizes ∏˜ to match features with similar pairwise distances (red outlined boxes) whose m/z ratios are close. (D) The final step of GromovMatcher plots the retention times of features from both datasets against each other and fits a spline interpolation f^ weighted by the estimated coupling weights ∏^ . This retention time drift function is then used to set all entries ∏˜ij to zero for those outlier pairs (fxi, fyj) which exceed twice the median absolute deviation (MAD) around f^ (green highlighted region). Finally, the coupling matrix ∏˜ is filtered and/or thresholded to obtain a refined coupling ∏^ which is then binarized to obtain a one-to-one matching M between a subset of metabolite pairs in both datasets.

OT was originally developed to optimize the transportation of soil for the construction of forts Monge (1781) and was later generalized through the language of probability theory and linear programming Kantorovich (2006), leading to efficient numerical algorithms and direct applications to planning problems in economics. The ability of OT to efficiently match source to target locations found applications in data science for the alignment of distributions Courty et al. (2017); Alvarez-Melis et al. (2019) and was generalized by the Gromov-Wasserstein (GW) method Peyré et al. (2016); Alvarez-Melis and Jaakkola (2018) to align datasets with features of differing dimensions.

In practice, a sizeable fraction of the metabolic features measured in one study may not be present in the other. Hence, in most cases only a subset of features in both datasets can be matched. Recent GW formulations for unbalanced matching problems Sejourne et al. (2021) allow for matching only subsets of metabolic features with similar intensity structures (Fig. 1c). To incorporate additional feature information, we modify the optimization objective of unbalanced GW to penalize feature matches whose m/z differences exceed a fixed threshold (Methods, Appendix 1). The optimization ofthis objective computes a coupling matrix ∏˜ where each entry ∏˜ij indicates the level of confidence in matching metabolic feature fxi in Dataset 1 to fyj in Dataset 2.

Differences in experimental conditions can induce variations in RT between datasets that can be nonlinear and large in magnitude Zhou et al. (2012); Climaco Pinto et al. (2022); Habra et al. (2021). In the spirit of previous methods for LC-MS batch or dataset alignment Smith et al. (2006); Brunius et al. (2016); Liu et al. (2020); Vaughan et al. (2012); Habra et al. (2021); Climaco Pinto et al. (2022); Skoraczynski et al. (2022), the learned coupling ∏˜ is used to estimate a nonlinear map (drift function) between RTs of both datasets by weighted spline regression, which allows us to filter unlikely matches from the coupling matrix to obtain a refined coupling matrix ∏^ (Fig. 1d, Methods). An optional thresholding step removes matches with small weights from the coupling matrix. The final output of GromovMatcher is a binary matching matrix M where Mij is equal to 1 if features fxi and fyj are matched and 0 otherwise. Throughout the paper, we refer to the two variants of GromovMatcher, with and without the optional thresholding step as GMT and GM respectively.

Validation on ground-truth data

We first evaluate the performance of GromovMatcher using a real-world untargeted metabolomics study of cord blood across 499 newborns containing 4,712 metabolic features characterized by their m/z, RT, and feature intensities Alfano et al. (2020). To generate ground-truth data, we randomly divide the initial dataset into two smaller datasets sharing a subset of features (Fig. 2). We simulate diverse acquisition conditions by adding noise to the m/z and RT of dataset 2, and to the feature intensities in both datasets. Moreover, we introduce an RT drift in dataset 2 to replicate the retention time variations observed in real LC-MS experiments (Methods and Materials). For comparison, we also test M2S Climaco Pinto et al. (2022) and metabCombiner Habra et al. (2021), both of which use m / z, RT, and median or mean feature intensities to match features (Fig. 3). MetabCombiner is supplied with 100 known shared metabolic features to automatically set its hyperparameters, while M2S parameters are manually fine-tuned to optimize the F1-score in each scenario (Appendix 2). We assess the performance of GM, GMT, metabCombiner, and M2S across 20 randomly generated dataset pairs in terms of their precision (fraction of true matches among the detected matches) and recall/sensitivity (fraction of true matches detected) averaged across 20 dataset pairs.

To investigate how the number of shared features affects dataset alignment, we generate pairs of LC-MS datasets with low, medium, and high feature overlap (25%, 50%, and 75%), while maintaining a medium noise level (Methods). Here we find that GM and GMT generally outperform existing alignment methods, with a recall above 0.95 while metabCombiner and M2S tend to be less sensitive (Fig. 3b). All methods drop in precision as the feature overlap is decreased, with GM and GMT still maintaining an average precision above 0.8.

Simulated data for testing untargeted metabolomics alignment methods. (A) Initial LC-MS dataset taken from the EXPOsOMICS project with m/z, RT, and feature intensities of p = 4, 712 metabolites identified in cord blood across n = 499 newborns. (B) Newborns (rows) are split into two disjoint groups of sizes n1 = 249 and n2 = 250 respectively and metabolic features (columns) are split into two equal groups of size p1 = p2 with overlap ⌊λp⌋ where λ = 0.25 ,0.5 ,0.75 (Methods). Datasets are perturbed by additive noise of magnitude (σM, σRT, σFI) and a nonlinear drift f (X) is applied to the RTs of dataset 2. (C) The two resulting datasets share σ = 25%, 50%, or 75% of the original dataset’s metabolic features.

Next we evaluate all four methods at low, moderate, and high noise levels for pairs of datasets with 50% over-lap in their features (Methods). Our results show that GMT, GM, and M2S maintain an average recall above 0.89, while metabCombiner’s recall drops below 0.6 for high noise. At large noise levels, RT drift estimation becomes more challenging, leading to a higher rate of false matches between metabolites (lower precision) for all four methods (Fig. 3 — figure supplement 1). Nevertheless, GMT obtains a high average precision and recall of 0.86 and 0.92 respectively.

A notable difference between GM, metabCombiner, and M2S lies in their use of feature intensities. MetabCombiner expects that the mean feature intensity rankings are identical across studies, while M2S assumes that shared features have similar median intensities. In contrast, GM uses both the mean feature intensities and their variances and covariances. In practice, differences in experimental assays or study populations can lead to greater variation in feature intensities, making matchings based on these statistics less reliable. Centering and scaling the feature intensities to unit variance avoids potential biases arising from inconsistent feature intensity magnitudes, but preserves correlations that GM leverages.

Exploring this further, we test how sensitive all four methods are to centering and scaling of feature intensities. MetabCombiner and M2S are tuned using the same methodology as for noncentered and non-scaled data. For M2S, we match features solely based on their m/z and RT. In this experiment (Figure 3 — figure supplement 2), the absence of intensity magnitude information significantly affects metabCombiner’s performance and, to a lesser extent, M2S. GM and GMT still obtain accurate matchings, due to their use of correlation structures which are preserved under centering and scaling.

Application to EPIC data

Next, we apply GM, metabCombiner and M2S to align datasets from the European Prospective Investigation into Cancer and Nutrition (EPIC) cohort, a prospective study conducted across 23 European centers. EPIC comprises more than 500,000 participants who provided blood samples at recruitment Riboli et al. (2002). Untargeted metabolomics data were successively acquired in several studies nested within the full cohort.

In the present work, we use LC-MS data from the EPIC cross-sectional (CS) study Slimani et al. (2003) and two matched case-control studies nested within EPIC, on hepatocellular carcinoma (HCC) Stepien et al. (2016, 2021) and pancreatic cancer (PC) Gasull et al. (2019). LC-MS untargeted metabolomic data were acquired at the International Agency for Research on Cancer, making use of the same platform and methodology (Methods). The number of samples and features in each study is displayed in Fig. 4a.

Comparison of MetabCombiner, M2S, and GromovMatcher on simulated data. (A) Ground-truth matchings, and matchings inferred by metabCombiner, M2S, GM, and GMT. Pairs of datasets are generated for three levels of overlap (low, medium and high), with a medium noise level (Methods). Matches correctly recovered (true positives) are represented in green. True matches that are not recovered (false negatives) are highlighted in grey. Incorrect matches (false positives) are plotted in red. (B) Average precision and recall on 20 randomly generated pairs of datasets, for three levels of overlap (low, medium and high) with a medium noise level.

Figure 3—figure supplement 1. Average precision and recall obtained on simulated data, with fixed overlap λ = 0.5. The noise level corresponds to different values of σRT and σFI. High, medium and low noise level correspond to (σRT, σFI) = (0.2, 0.1), (0.5, 0.5) and (1, 1) respectively. We run 20 simulations for each setting.

Figure 3—figure supplement 2. Performance on centered and scaled data. The feature intensities of both datasets are centered and scaled to have means of0 and standard deviations of 1. The average precision and recall of the three methods are computed on 20 randomly generated pairs of datasets, for (a) three levels of overlap (low, medium and high corresponding to λ = 0.25 ,0.5 and 0.75, respectively) with a medium noise level ((σRT, σFI) = (0.5, 0.5)), and (b) fixed medium overlap (λ = 0.5) and three different noise levels (low, medium and high corresponding to (σRT, σFI) = (0.2, 0.1), (0.5, 0.5) and (1, 1), respectively).

Loftfield et al. Loftfield et al. (2021) previously matched features from the CS, HCC, and PC studies in EPIC for alcohol biomarker discovery. The authors first identified 205 features (163 in positive and 42 in negative mode) associated with alcohol intake in the CS study. These features were then manually matched by an expert to features in both the HCC and PC studies (Methods). In our analysis, we use these features as a validation set and compare each method’s matchings to the expert manual matchings on this subset. Due to the imbalance between the number of positive and negative mode features in the validation subset, our main analysis focuses on the alignment results of CS with HCC and CS with PC in positive mode. We delegate the matching results between the negative mode studies to Appendix 3.

In this section, we use the same settings for GM as in our simulation study, and do not apply an additional thresholding step. The parameters of metabCombiner and M2S are calibrated using the validation subset as prior knowledge (Appendix 2).

Preliminary analysis of the validation subset reveals inconsistencies in the mean feature intensities (Figure 4 — figure supplement 1), but Figure 4b shows that on centered and scaled data, the 90 expert matched features shared between the CS and HCC studies have similar correlation structures. Hence, to avoid potential errors we center and scale the feature intensities which improves the performance of all three methods tested below (Appendix 3, Appendix 3 — Table 1).

Hepatocellular carcinoma

Here we analyze the quality of the matchings obtained by GM, M2S, and metabCombiner between the CS and HCC datasets in positive mode. Both GM and M2S identify approximately 1000 shared features while metabCombiner finds a smaller number of about 700 shared features. We refer the reader to Figure 4- figure supplement 2a for the precise matched feature sizes and details on the agreement between the feature matchings of all three methods.

We evaluate the performance of metabCombiner, M2S, and GM on the validation subset in positive mode (Fig. 4c), which consist of 90 features from the CS study manually matched to features from the HCC study and 73 features specific to the CS study. MetabCombiner demonstrates precise matching but lacks sensitivity. M2S’s precision and recall are comparable with GM, in contrast to its performance on simulated data. This can be attributed to the RT drift shape between the CS and HCC studies (Appendix 2), which is estimated to be close to linear (Figure 4 — figure supplement 3). Because the parameters of M2S are fine-tuned in the validation subset, it is able to learn this linear drift and apply tight RT thresholds to achieve accurate matchings. In contrast to metabCombiner and M2S, the GM algorithm is not given any prior knowledge of the validation subset, and nevertheless demonstrates the highest precision and recall rates of the three methods (Fig. 4c). Figure 4b shows how GM recovers the majority of the expert matched pairs by leveraging the shared correlations.

Pancreatic cancer

Matching features between the CS and PC studies in positive mode, GM and M2S identify approximately 1000 common features, while metabCombiner detects approximately 600 matches (Figure 4- figure supplement 2b). We examine the performance of all three methods on the validation subset consisting of 66 manually matched features between CS and PC along with 97 features specific to the CS study. As before, GM and M2S have high recall while the recall of metabCombiner is less than 0.5.

A decrease in precision is observed for both GM and M2S compared to the previous CS-HCC matchings. We therefore manually inspect the false positive matches; the set of CS features matched by the method to the PC study but explicitly examined and left unmatched in the expert manual matching. Assessing the GM results, we identify 7 false positive feature matches. Upon secondary inspection, 3 pairs are revealed as correct matches that were not initially identified in the expert matching. M2S finds 11 false positive matches which include the 7 false positives recovered by GM. Manual examination of the 4 remaining pairs reveals 2 clear mismatches. These results highlight the advantage of using automated methods for data alignment, as both GM and M2S detect correct matches that were not identified by experts, with GM being more precise than M2S.

Application of GromovMatcher and comparison to existing methods on EPIC dataset. (A) Dimensions of the three EPIC studies used. For each ionization mode, the cross-sectional (CS) study is aligned successively with the hepatocellular carcinoma (HCC) study and the pancreatic cancer (PC) study. (B) Demonstration of expert manual matching and GromovMatcher (GM) matching between the CS and HCC studies in positive mode. Experts manually match 90 features from Loftfield et al. (2021) and the correlation matrices of these features in both datasets have similar structure (bottom two matrices). GM discovers 996 shared features between the CS and HCC datasets which have similar correlation structure (top two matrices). We validate that 88 of the 90 features from the manually expert matched subset are contained in the set of features matched by GM. (C) Performance of metabCombiner (mC), M2S and GM in positive mode. Precisions and recalls are measured on a validation subset of 163 manually examined features, and 95% confidence intervals are computed using modified Wilson score intervals. (D) Performance of mC, M2S and GM in negative mode. Precision and recall are measured on a validation subset of42 manually examined features, and 95% confidence intervals are computed using modified Wilson score intervals. See Table 2 and Table 3 for exact precisions, recalls, and confidence intervals in positive and negative mode respectively.

Figure 4—figure supplement 1. Consistency of the mean feature intensities (FI) in EPIC. Each scatter plot represents the mean feature intensities of manually matched features from the validation subset. Each dot represents a pair of manually matched features. The axis represent the mean feature intensities recorded in the two different studies.

Figure 4—figure supplement 2. Overlap between the matching results obtained by metabCombiner, M2S and GromovMatcher in EPIC. Venn diagrams are not up to scale.

Figure 4—figure supplement 3. Estimated RT drift between the EPIC studies aligned in the main experiment. Each dot correspond to a candidate matched pair after the first step of GM (m/z constrained GW matching), before the RT drift estimation and RT based filtering.

Illustration for alcohol biomarker discovery

Loftfield et al. Loftfield et al. (2021) identified biomarkers of habitual alcohol intake by first performing a discovery step, where they examined the relationship between alcohol intake and metabolic features in the CS study. They then manually matched the significant features in CS to features from the HCC and PC studies, and repeated the analysis with samples from the HCC and PC studies to determine whether the association with alcohol intake persisted. This led to the identification of 10 features possibly associated with alcohol intake (Fig. 5a).

To extend this analysis and illustrate the benefit of GM automatic matching for biomarker discovery, we use GM to pool features from the CS, HCC, and PC studies, and examine the relationship between metabolic features and alcohol intake in the pooled study (Methods and Fig. 5b).

Applying an FDR correction on the pooled study, we identify 243 features associated with alcohol intake, including 185 features consistent with the discovery step of Loftfield et al. Loftfield et al. (2021), and 55 newly discovered features (Fig. 5c). Using the more stringent Bonferroni correction on the pooled data, we identify 36 features shared by all three studies that are significantly associated with alcohol intake. These features include all 10 features identified in Loftfield et al. (Fig. 5c). These findings highlight the potential benefits of using GM automatic matching for biomarker discovery in untargeted metabolomics data. Additional information regarding the methodology and findings of our GM and Loftfield et al. analyses can be found in Methods and Appendix 3.

Discussion

LC-MS metabolomics has emerged as an increasingly powerful tool for biological and biomedical research, offering promising opportunities for epidemiological and clinical investigations. However, integrating data from different sources remains challenging. To address this issue, we introduce GromovMatcher, a method based on optimal transport that automatically aligns LC-MS data from pairs of studies. Our method exhibits superior performance on both simulated and real data when compared to existing approaches. Additionally, it presents a user-friendly interface with few hyperparameters.

While GromovMatcher is robust to noise and variations in data, it may face limitations when aligning LC-MS studies from populations with different characteristics, where the correlation structures between features may be inconsistent across studies. In this case, the base assumption of GromovMatcher can be relaxed by focusing on subsamples with similar characteristics, as exemplified in a recent study Gomari et al. (2022).

A current limitation is that GromovMatcher does not account for more than two datasets simultaneously, although this can be overcome by aligning multiple studies to a chosen reference dataset, as demonstrated in our biomarker experiments. The extension of Gromov-Wasserstein to multiple distributions Beier et al. (2022) is another promising approach for generalizing Gro-movMatcher to multiple dataset alignment. Further improvements can be made by incorporating existing knowledge about the studies being matched, such as known shared features, samples in common, or MS/MS data.

The results obtained from GromovMatcher are highly promising, opening the door for various analyses of metabolomic datasets acquired in different experimental laboratories. Here we demonstrated the potential of GromovMatcher in expediting the combination and meta-analysis of data for biomarker and metabolic signature discovery. The matchings learned by GromovMatcher also allow for comparison between experimental protocols by assessing the drift in m/z, RT, and feature intensities across studies. Finally, inter-institutional annotation efforts can directly benefit from incorporating this method to transfer annotations between aligned datasets. Bridging the gap between otherwise incompatible LC-MS data, GromovMatcher enables seamless comparison of untargeted metabolomics experiments.

Comparison of GromovMatcher and Loftfield et al. Loftfield et al. (2021) analysis for alcohol biomarker discovery on EPIC data. (A) Loftfield study implemented a discovery step, examining the relationship between alcohol intake and metabolic features in the CS study. The significant features in CS were manually matched to features from the HCC and PC and the analysis was repeated using samples from the HCC and PC studies. After this step, 10 features associated with alcohol intake were identified. (B) GromovMatcher analysis begins by matching features from CS study to HCC and PC studies respectively (top blue, yellow, and red boxes). Samples corresponding to each CS feature are combined with the samples of its matched feature in the HCC study, PC study, or both. This generates a larger pooled data matrix with the same number of features as the CS study but with more samples pooled across the three original studies (center matrix). Because some features in the CS study may not have matches in HCC or PC, the corresponding entries in the pooled matrix are set to NaN/missing values (white regions in matrix). Each column/feature in this matrix is statistically tested for association with alcohol intake (ignoring missing values) and an FDR or a stricter Bonferroni correction is performed to retain only a subset of features from the pooled study that have a strong association. (C) Venn diagrams show intersection of feature sets (in positive and negative mode) found to be associated with alcohol intake by one of the four different analyses.

Methods and Materials

GromovMatcher method overview

GromovMatcher accepts as input two feature tables from separate LC-MS untargeted metabolomics studies. Each feature table for dataset 1 and dataset 2 consists of n1, n2 biospecimen samples respectively and p1, p2 metabolic features respectively detected in the study. Features in dataset 1 are given the label fxi for i = 1,… ,p1. Every feature is characterized by a mass-to-charge ratio (m/z) denoted by mix , a retention time (RT) denoted by RTix , and a vector of intensities across all samples written as Xi∈ℝn1 . Similarly, features in dataset 2 are labeled as fyj for j = 1,… ,p2 and are characterized by their m/z mjy , retention time RTjy , and a vector of intensities across all samples Yi∈ℝn2 .

Our goal is to identify pairs of indexes (i, j) with i ∈ {1, … , p1} and j ∈ {1, …, p2}, such that fxi and fyj correspond to the same metabolic feature. More formally, we aim to identify a matching matrix M*∈{0,1}p1×p2 such that Mij = 1 if fxi and fyj correspond to the same feature, hereafter referred to as matched features. Otherwise, we set Mij = 0.

Because the m/z and RT values of metabolomic features are often noisy and subject to experimental bias, our matching algorithm leverages metabolite feature intensities Xi,Yj to produce accurate dataset alignments. The GromovMatcher method is based on the idea that signal intensities of the same metabolites measured in two different studies should exhibit similar correlation structures, in addition to having compatible m/z and RT values. Here we define the Pearson correlation for vectors u,v ∈ ℝn as corr(u,v)=〈u,u¯,v−v¯〉‖u−u¯‖‖v−v¯‖ (1)

where we define u¯=1n∑i=1nui,‖u‖=∑i=1nui2,〈u,v〉=∑i=1nuivi (2)

as the mean value, Euclidean norm and inner product respectively. If measurements Xi,Yj correspond to the same underlying feature, and similarly, measurements Xk,Yt share the same an underlying feature, we expect that corr(Xi,Xk)≈corr(Yj,Yi). (3)

This idea that the feature intensities of shared metabolites have the same correlation structure in both datasets also holds more generally for distances, under a suitable choice of distance. For example, the correlation coefficient corr(u, V) can be turned into a dissimilarity metric by defining dcos(u,v)=1−corr(u,v) (4)

commonly referred to as the cosine distance. Preservation of feature intensity correlations then trivially amounts to the preservation of cosine distances.

Another classical notion of distance between vectors u, V ∈ ℝn is the normalized Euclidean distance deuc(u,v)=1n‖u−v‖=1n∑i=1n(ui−vi)2 (5)

which is equal to the cosine distance (up to constants) when the vectors u, V are centered and scaled to have zero mean and a standard deviation of one. The Euclidean distance depends on the magnitude or mean intensity of metabolic features, and hence is a useful metric for matching metabolites as long as these mean feature intensities are reliably collected.

To summarize, the main tenant of GromovMatcher is that if measurements Xi,Yj correspond to the same feature and Xk, Yl correspond to the same feature, then for suitably chosen distances dx:ℝn1×ℝn1→ℝ and dy:ℝn2×ℝn2→ℝ , these distances are preserved dx(Xi,Xk)≈dy(Yj,Yl) (6)

across both datasets. In this paper, the distances dx,dy are taken to be the normalized Euclidean distances in (5). We take care to specify those experiments where the metabolic features X and Y are centered and scaled. In these cases, implicitly the Euclidean distance between normalized feature vectors becomes the cosine distance (4) between the original (unnormalized) feature vectors.

Unbalanced Gromov-Wasserstein

The goal ofGromovMatcher is to learn a matching matrix M∈{0,1}p1×p2 that gives an alignment between a subset of metabolites in both datasets. However, searching over the combinatorially large set of binary matrices would be an inefficient approach for dataset alignment. The mathematical framework of optimal transport Peyré et al. (2019) instead enlarges this space of binary matrices to the set of coupling matrices with real nonnegative entries ∏∈ℝ+p1×p2 . The entries ∏ij with large weights indicate that feature fxi in dataset 1 and feature fyj in dataset 2 are a likely match. Taking inspiration from (6), we minimize the following objective function ε(Π)∑i,k=1p1∑i,l=1p2ΠijΠkl|dx(Xi,Xk)−dy(Yj,Yl)| (7)

to estimate the coupling matrix ∏.

A standard approach is to optimize this objective over all coupling matrices ∏ under exact marginal constraints ∏1p2=1p11p1,∏T1p1=1p21p2 . Here we define 1n is the ones vector of length n, and ∏1=∏1p2,∏2=∏T1p1 denote the column and row sums of the coupling matrix. Objective (7) under these exact marginal constraints defines a distance between the two sets of metabolic feature vectors {Xi}i=1p1,{Yi}i=1p2 known as the Gromov-Wasserstein distance Mémoli (2011), a generalization of optimal transport to metric spaces. Note that for pairs Xi,Yj and Xk,Yl for which dx(Xi, Xk) ≈ dy(Yj,Yl), the entries ∏ij, ∏kl are penalized less and hence matches between features fxi ,fyj and features fxk ,fyl are more favored. In our optimization, we avoid enforcing exact marginal constraints on the marginal distributions ∏1p2 and ∏T1p1 of our coupling matrix as this would enforce that all metabolites in both datasets are matched (Appendix 1). However, without any marginal constraints on the coupling ∏, the objective function (7) is trivially minimized by ∏ = 0, leaving all metabolites in both datasets unmatched.

To account for this, we follow the ideas of unbalanced Gromov-Wasserstein (UGW) Sejourne et al. (2021) and add three regularization terms to our objective Lρ,ε(Π)=ε(Π)+ρDKL(Π1⊗Π1,a⊗a) +ρDKL(Π2⊗Π2,b⊗b) +εDKL(Π⊗Π,(a⊗a)⊗2) (8)

where ρ, ε > 0 and we define a = 1p1,b = 1p2. Here ⊗ denotes the Kronecker product. We define DKL as the Kullback-Leibler (KL) divergence between two discrete distributions μ,ν∈ℝ+p by DKL(μ,v)=∑i=1pμi In (μivi)−∑i=1pμi +∑i=1pvi (9)

which measures the closeness of probability distributions.

The first two regularization terms in (8) enforce that the row sums and column sums of the coupling matrix ∏ do not deviate too much from a uniform distribution, leading our optimization to match as many metabolic features as possible. The magnitude of the regularizer ρ roughly enforces the fraction of metabolites in both datasets that are matched where large ρ implies most metabolites are matched across datasets. The final regularization term ε in (8) controls the smoothness (entropy) of the coupling matrix ∏ where larger values of ε encourage ∏ to put uniform weights on many of its entries, leading to less precision in the metabolite matches. However, increasing ε also leads to better numerical stability and a significant speedup of the alternating minimization algorithm used to optimize the objective function (Appendix 1). In our implementation, we set ρ and ε to the lowest possible values under which our optimization converges, with ρ = 0.05 and ε = 0.005.

Our full optimization problem can now be written as UGWρ,ε=Π∈ℝ+p1×p2minLρ,ε(Π). (10)

The UGW objective function is optimized through alternating minimization based on the code of Sejourne et al. (2021) using the unbalanced Sinkhorn algorithm Séjourné et al. (2019) from optimal transport (Appendix 1).

Constraint on m/z ratios

Matched metabolic features must have compatible m/z so we enforce that ∏ij = 0 when |mix−mjy|>mgap mgap where mgap is a user-specified threshold. Based on prior literature Loftfield et al. (2021); Hsu et al. (2019); Climaco Pinto et al. (2022); Habra et al. (2021); Chen et al. (2021) we set mgap = 0.01ppm. Note that mgap is not explicitly used in (10) but is rather enforced in each iteration of our alternating minimization algorithm for the UGW objective (Appendix 1).

Unlike the m/z ratios discussed above, RTs often exhibit a non-linear deviation (drift) between studies so we cannot enforce compatibility of RTs directly in our optimization. Instead, in the following step of our pipeline we ensure matched metabolite pairs have compatible RTs by estimating the drift function and subsequently using it to filter out metabolite matches whose RT values are inconsistent with the estimated drift.

Estimation of the RT drift and filtering

Estimating the drift between RTs of two studies is a crucial step in assessing the validity of metabolite matches and discarding those pairs which are incompatible with the estimated drift.

Let ∏˜∈ℝ+p1×p2 be the minimizer of (10) obtained after optimization. We seek to estimate the RT drift function f : ℝ+ → ℝ+ which relates the retention times of matched features between the two studies. Namely, if feature fxi and feature fyj correspond to the same metabolic feature, then we must have that RTjy≈f(RTix) .

We propose to learn the drift f through the weighted spline regression f∈Bn,kmin∑i=1p1∑j=1p2Π˜ij|f(RTix)−RTjy| (11)

where Bn,k is the set of n-order B-splines with k knots. All pairs RTix,RTjy in objective (11) are weighted by the coefficients of ∏˜ so that larger weights are given to pairs identified with high confidence in the first step of our procedure. The order of the B-splines was set to n = 3 by default, while the number of knots k was selected by 10-fold cross-validation.

Pairs identified as incompatible with the estimated RT drift are then discarded from the coupling matrix. To do this, we first take the estimated RT drift f^ , and the set of pairs S={i,j:∏˜i,j≠0} recovered in ∏˜ . We then define the residual associated with (i,j) ∈ S as rf^(i,j)=|f^(RTix)−RTjy|. (12)

The 95% prediction interval and the median absolute deviation (MAD) of these residuals are given by PI=1.96×std({rf^(i,j),(i,j)∈S})MAD=median({|rf^(i,j)−μr|,(i,j)∈S})μr=median({|rf^(i,j)|,(i,j)∈S}) (13)

where |S| is the size of S and the functions std and median denote the standard deviation and median respectively. Similar to the approach in Climaco Pinto et al. (2022), we create a new filtered coupling matrix ∏^∈ℝ+p1×p1 given by Π^ij={Π~ijif rf^(i,j)<μr+rthresh0otherwise. (14)

where rthresh is a given filtering threshold. Following Habra et al. (2021), the estimation and outlier detection step can be repeated for multiple iterations, to remove pairs that deviate significantly from the estimated drift and improve the robustness of the drift estimation. In our main algorithm, we use two preliminary iterations where estimate the RT drift and discard outliers outside of the 95% prediction interval by setting rthresh = PI. We the re-estimate the drift and perform a final filtering step with the more stringent MAD by setting rthresh = 2 × MAD.

At this stage, it is possible for ∏^ to still contain coefficients of very small magnitude. As an optional postprocessing step, we discard these coefficients by setting all entries smaller than τmax(∏^) to zero, for some user-defined τ ∈ [0,1]. Lastly, a feature from either study could have multiple possible matches, since ∏^ can have more than one non-zero coefficient per row or column. Although reporting multiple matches can be helpful in an exploratory context, for the sake of simplicity in our analysis, the final output of GromovMatcher returns a one-to-one matching, as we only keep those metabolite pairs (i, j) where the entry ∏^ij is largest in its corresponding row and column. All nonzero entries of ∏^ which do not satisfy this criterion are set to zero. Finally, we convert ∏^ into a binary matching matrix M ∈ {0, 1}p1 × p2 with ones in place of its nonzero entries and this final output is returned to the user.

As a naming convention, we use the abbreviation GM for our GromovMatcher method, and use the abbreviation GMT when running GromovMatcher with the optional τ-thresholding step with τ = 0.3.

Metrics for dataset alignment

Every alignment method studied in this paper returns a binary partial matching matrix M ∈ {0, 1} p1×p1 which has at most one nonzero entry in each row and column. Specifically, Mij = 1 if metabolic features i and j in both datasets correspond to each other and Mij = 0 otherwise. In our simulated experiments, we compare the partial matching M to a known ground-truth partial matching matrix M* ∈ {0, 1}p1×p2.

To do this, we first compute the number of true positives, false positives, true negatives, and false negatives as TP=∑i=1p1∑j=1p21Mij=11Mij*=1FP=∑i=1p1∑j=1p21Mij=11Mij*=0TN=∑i=1p1∑j=1p21Mij=01Mij*=0FN=∑i=1p1∑j=1p21Mij=01Mij*=1 (15)

where 1 denotes the indicator function. Then we use these values to compute the precision and recall as Precision=TPTP+FPRecall=TPTP+FN. (16)

Precision measures the fraction of correctly found matches out of all discovered metabolite matches, while recall, also know as sensitivity, measures the fraction of correctly matched pairs out of all truly matched pairs. These two statistics can be summarized into one metric called the F1-score by taking their harmonic mean F1=2⋅Precision⋅RecallPrecision+Recall (17)

These three metrics, precision, recall, and the F1-score, are used throughout the paper to assess the performance of dataset alignment methods, both on simulated data where the ground-truth matching is known, and on the validation subset in EPIC, using results from the manual examination as the ground-truth benchmark.

Validation on simulated data

To assess the performance of GromovMatcher and compare it to existing dataset alignment methods, we simulate realistic pairs of untargeted metabolomics feature with known ground-truth matchings. This allows us to analyze the dependence of alignment methods on the number of shared metabolites, dataset noise level, and feature intensity centering and scaling.

Dataset generation

Our pairs of synthetic feature tables are generated from one real untargeted metabolomics study of 500 newborns within the EXPOsOMICS project, which uses reversed phase liquid chromatographyquadrupole time-of-flight mass spectrometry (UHPLC-QTOF-MS) system in positive ion mode Alfano et al. (2020). The original dataset is first preprocessed following the procedure detailed in Alfano et al. Alfano et al. (2020), resulting in p = 4,712 features measured in n = 499 samples available for subsequent analysis. Features and samples from the original study are then divided into two feature tables of respective size (n1, p1) and (n2,p2), with n1 + n2 = n and p1, p2 ≤ p. In order to do this, n1=⌈n/2⌉ randomly chosen samples from the original study are placed into dataset 1 and the remaining n2=⌊n/2⌋ samples from the original study are placed into dataset 2. Here ⌊⋅⌋ and ⌈⋅⌉ denote integer floor and ceiling functions. The features of the original study are randomly assigned to dataset 1, dataset 2, or both, allowing the resulting studies to have both common and studyspecific features (Fig. 2). Specifically, for a fixed overlap parameter λ ∈ [0,1], we assign a random subset ⌊λp⌋ of the features into both dataset 1 and dataset 2 while the remaining p−⌊λp⌋ features are divided equally between the two studies such that p1 = p2. We choose λ ∈ {0.25, 0.5,0.75} corresponding to low, medium and high overlap.

After generating a pair of studies, random noise is added to the m/z, RT and intensity levels of features in dataset 2 to mimic variations in data acquisition across two different experiments. The noise added to each m / z value in study 2 is sampled from a uniform distribution on the interval [-σM, σM] with σM = 0.01 Climaco Pinto et al. (2022). The RTs of dataset 2 are first deviated by the function f(x)=1.1x+1.3sin(1.2x) , corresponding to a systematic inter-dataset drift Habra et al. (2021); Climaco Pinto et al. (2022); Brunius et al. (2016). A uniformly distributed noise on the interval [-σRT, σRT] is added to the deviated RTs of dataset 2, with σRT ∈ {0.2, 0.5, 1} (in minutes) corresponding to low, moderate and high variations Climaco Pinto et al. (2022); Habra et al. (2021); Vaughan et al. (2012). Finally, we add a Gaussian noise N(0,σFI2) to the feature intensities of both studies where σFI is the scalarvariance ofthe noise. This noise perturbs the correlation matrices of dataset 1 and dataset 2, making matching based on feature intensity correlations more challenging. We vary σFI over the set of values {0.1, 0.5, 1}.

Given this data generation process, we test the performance of the four alignment methods (M2S, metabCombiner, GM, and GMT) under the parameter settings described below.

Dependence on overlap

We first assess how the performance of the four methods is affected by the number of metabolic features shared in both datasets. For each value of λ = 0.25 ,0.5 ,0.75 (low, medium and high overlap), we randomly generate 20 pairs of datasets with noise on the m/z , RT and feature intensities set to σM = 0.01, σRT = 0.5, σFI = 0.5. The precision and recall of each method at low, medium, and high overlap is recorded for each of the repetitions.

Noise robustness

Next we test the robustness to noise of each method by fixing the metabolite overlap fraction at λ = 0.5 and generating 20 random pairs of datasets at low (σRT = 0.2, σFI = 0.1), medium (σRT = 0.5, σFI = 0.5), and high (σRT = 1, σFI = 1) noise levels. Similarly, the precision and recall of each method is saved for each noise level across the 20 repetitions.

Feature intensity centering and scaling

In order to test how all four methods are affected when the mean feature intensities and variance are not comparable across studies, we assess their performance when the feature intensities in both studies are mean centered and standardized to have unit standard deviation across all samples. We again generate 20 random pairs of datasets with medium overlap and medium noise, normalize the feature intensities in each pair of datasets, and compute the precision and recall of each method across the 20 repetitions.

EPIC data

We also evaluate our method on data collected within the European Prospective Investigation into Cancer and Nutrition (EPIC) cohort, an ongoing multicentric prospective study with over 500,000 participants recruited between 1992 and 2000 from 23 centers in 10 European countries, and who provided blood samples at the inclusion in the study Riboli et al. (2002). In EPIC, untargeted metabolomics data were successively acquired in several studies nested within the full cohort.

In the present work, we use untargeted metabolomics data acquired in three studies nested in EPIC, namely the EPIC cross-sectional (CS) study Slimani et al. (2003) and two matched case-control studies nested within EPIC, on hepatocellular carcinoma (HCC) Stepien et al. (2016, 2021) and pancreatic cancer (PC) Gasull et al. (2019), respectively. All data were acquired at the International Agency for Research on Cancer, making use of the same plateform and methodology: UHPLC-QTOF-MS (1290 Binary Liquid chromatography system, 6550 quadrupole time-of-flight mass spectrometer, Agilent Technologies, Santa Clara, CA) using reversed phase chromatography and electrospray ionization in both positive and negative ionization mode.

In a previous analysis aiming at identifying biomarkers of habitual alcohol intake in EPIC, the 205 features associated with alcohol intake in the CS study were manually matched to features in both the HCC and PC studies Loftfield et al. (2021). The results from this manual matching are presented in Table 1.This matching process was based on the proximity of m/z and RT, using a matching tolerance of ±15 ppm and ±0.2 min, and on the comparison of the chromatograms of features in a quality control samples from both studies.

Preprocessing

In the HCC and PC studies, samples corresponding to participants selected as cases in either study (i.e., participants selected in the study because ofa diagnosis of incident HCC or PC) are excluded. Indeed, the metabolic profiles of participants selected as controls are expected to be more comparable across studies than those of cases, especially if certain features are associated with the risk of HCC or PC. Apart from this additional exclusion criterion, the untargeted metabolomics data of each study is pre-processed following the steps described in Loftfield et al. Loftfield et al. (2021), to eliminate unreliable features and samples, impute missing values and minimize technical variations in the feature intensity levels.

Alcohol biomarker discovery

Loftfield et al. Loftfield et al. (2021) used the untargeted metabolomics data of the CS, HCC and PC studies in their alcohol biomarker discovery study in EPIC, without being able to automatically match their common features and pool the 3 datasets. Instead, the authors first implemented a discovery step, examining the relationship between alcohol intake and metabolic features measured in the CS study and accounting for multiple testing using a false discovery rate (FDR) correction. This led to the identification of 205 features significantly associated with alcohol intake in the CS study. In order to gauge the robustness of these associations, the authors of Loftfield et al. Loftfield et al. (2021) then implemented a validation step using data from two independent test sets. The first test set was composed of data from the EPIC HCC and PC studies, while the second was derived from the Finnish Alpha-Tocopherol, Beta-Carotene Cancer Prevention (ATBC) study. The 205 features identified in the discovery step were manually investigated for matches in the EPIC test set, and 67 features were effectively matched to features in the HCC or PC study, or both. The authors then evaluated the association between alcohol intake and those 67 features, applying a more conservative Bonferroni correction to determine whether the association with alcohol intake persisted. This step led to the identification of 10 features associated with alcohol intake (Extended Data Fig. 5a). The second test set was then used to determine whether those 10 features were also significant in the ATBC population, which was indeed the case.

To conduct a more in-depth investigation of the matchings produced by the GromovMatcher algorithm, we build upon the analysis previously conducted by Loftfield et al. Loftfield et al. (2021) by exploring potential alcohol biomarkers using a pooled dataset created from the CS, HCC, and PC studies. Our goal is to assess whether pooling the data leads to increased statistical power and allows for the detection of more features associated with alcohol intake. Namely, we generate the pooled dataset by aligning a chosen reference dataset (CS study) with the HCC and PC studies successively using the GM matchings computed in both positive and negative mode (Methods and Extended Data Fig. 5b). Features that are not detected in either the HCC or PC studies are designated as ‘missing’ in the final pooled dataset for samples belonging to the respective studies where the feature is not found.

To evaluate the potential relationship between alcohol consumption and pooled metabolic features, we use a methodology akin to that of Loftfield et al. Loftfield et al. (2021). The self-reported alcohol intake data is adjusted for various demographic and lifestyle factors (age, sex, country, body-mass-index, smoking status and intensity, coffee consumption, and study) via the residual method in linear regression models. Feature intensities are also adjusted for technical variables (plate number and position within the plate) via linear mixed effect models. The significance of the association is assessed using correlation coefficients computed from the residuals for both selfreported alcohol intake and feature intensities. P-values are corrected using either false discovery rate (FDR) or Bonferroni correction to account for multiple testing. Corrected p-values less than 5% are considered significant.

eLife Assessment

This paper represents important findings when identifying untargeted metabolomics and its differences between metabolomes of different biological samples. GromovMatcher is the fantasy name for the soft development. The main idea behind it is built on the assumption of featuring and matching complex datasets. Although the manuscript reflects a solid analysis, it remains incomplete for validation with putative non-curated datasets.

Reviewer #1 (Public Review):

Summary:

The authors have implemented the Optimal Transport algorithm in GromovMatcher for comparing LC/MS features from different datasets. This paper gains significance in the proteomics field for performing meta-analysis of LC/MS data.

Strengths:

The main strength is that GromovMatcher achieves significant performance metrics compared to other existing methods. The authors have done extensive comparisons to claim that GromovMatcher performs well.

Weaknesses:

There are two weaknesses.

1. When the number of features is reduced the precision drops to ~0.8.

2. How applicable is the method for other non-human datasets?

Reviewer #2 (Public Review):

Summary:

The goal of untargeted metabolomics is to identify differences between metabolomes of different biological samples. Untargeted metabolomics identifies features with specific mass-to-charge ratio (m/z) and retention time (RT). Matching those to specific metabolites based on the model compounds from databases is laborious and not always possible, which is why methods for comparing samples on the level of unmatched features are crucial.

The main purpose of the GromovMatcher method presented here is to merge and compare untargeted metabolomes from different experiments. These larger datasets could then be used to advance biological analyses, for example, for the identification of metabolic disease markers. The main problem that complicates merging different experiments is m/z and RT vary slightly for the same feature (metabolite).

The main idea behind the GromovMatcher is built on the assumption that if two features match between two datasets (that feature i from dataset 1 matches feature j from dataset 2, and feature k from dataset 1 matches feature l from dataset 2), then the correlations or distances between the two features within each of the datasets (i and k, and j and l) will be similar. The authors then use the Gromov-Wasserstein method to find the best matches matrix from these data.

The variation in m/z between the same features in different experiments is a user-defined value and it is initially set to 0.01 ppm. There is no clear limit for RT deviations, so the method estimates a non-linear deviation (drift) of RT between two studies. GromovMatcher estimates the drift between the two studies and then discards the matching pairs where the drift would deviate significantly from the estimate. It learns the drift from a weighted spline regression.

The authors validate the performance of their GromovMatcher method by a validation experiment using a dataset of cord blood. They use 20 different splits and compare the GromovMatcher (both its GM and GMT iterations, whereby the GMT version uses the deviation from estimated RT drift to filter the matching matrix) with two other matching methods: M2S and metabCombiner.

The second validation was done using a (scaled and centered) dataset of metabolics from cancer datasets from the EPIC cohort that was manually matched by an expert. This dataset was also used to show that using automatic methods can identify more features that are associated with a particular group of samples than what was found by manual matching. Specifically, the authors identify additional features connected to alcohol consumption.

Strengths:

I see the main strength of this work in its combination of all levels of information (m/z, RT, and higher-order information on correlations between features) and using each of the types of information in a way that is appropriate for the measure. The most innovative aspect is using the Gromov-Wasserstein method to match the features based on distance matrices.

The authors of the paper identify two main shortcomings with previously established methods that attempt to match features from different experiments: a) all other methods require fine-tuning of user-defined parameters, and, more importantly, b) do not consider correlations between features. The main strength of the GromovMatcher is that it incorporates the information on distances between the features (in addition to also using m/z and RT).

Weaknesses:

The first, minor, weakness I could identify is that there seem not to be plenty of manually curated datasets that could be used for validation. The second is also emphasized by the authors in the discussion. Namely, the method as it is set up now can be directly used only to compare two datasets.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

This paper represents important findings when identifying untargeted metabolomics and its differences between metabolomes of different biological samples. GromovMatcher is the fantasy name for the soft development. The main idea behind it is built on the assumption of featuring and matching complex datasets. Although the manuscript reflects a solid analysis, it remains incomplete for validation with putative non-curated datasets.

We are grateful to the eLife editor for taking the time and effort to assess our manuscript.

We are however unsure of what the editor means by “it remains incomplete for validation with putative non-curated datasets”. As noted by Reviewer 2, manually curated datasets that could be used for validation are scarce. Most publicly available datasets do not contain sufficient information to establish a ground truth matching on which GromovMatcher, M2S, or metabCombiner can be tested. Even in the case where such a ground truth matching can be established, it must be performed by-hand through a manual matching process which is extremely time-consuming and requires very specific expertise. This, in our opinion, only highlights the need for automatic alignment methods such as metabCombiner, M2S or GromovMatcher.

We do agree that the performance of GromovMatcher (and its competitors) needs to be validated further, and we plan to continue validating GromovMatcher as additional data becomes available in EPIC and other cohorts. With that in mind, the lack of publicly available validation data is the reason why we conducted such an extensive simulation study, arguably more comprehensive than previous validations, exploring challenging settings that we believe reflect real-life scenarios (main text “Validation on ground-truth data” and Appendix 3). We would like to stress that this allows us to highlight previously ignored limitations of the previously published methods, metabCombiner and M2S.

We wish to thank the editor and reviewers for their time and efforts in reviewing our manuscript which led to many significant additions to our paper. Namely we:

• Performed an additional sensitivity analysis (Appendix 3) exploring how an imbalance in the number of features or samples between two studies being matched (e.g. the dataset split), affects the quality of matchings found by GromovMatcher, metabCombiner, and M2S.

• Investigated how changing or removing the reference dataset (Appendix 5) in the EPIC study (main text “Application to EPIC data”), affects the results of GromovMatcher.

• Improved alignment matrix visualizations in Fig. 3a for all four methods tested on the validation data, to highlight more clearly which feature matches were correctly identified or missed.

The revised paper is uploaded as the file “main_elife_revision.pdf” where all revisions are highlighted in blue as well as a copy “main_elife_revision_nohighlights.pdf” where revisions are not highlighted.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

The authors have implemented the Optimal Transport algorithm in GromovMatcher for comparing LC/MS features from different datasets. This paper gains significance in the proteomics field for performing meta-analysis of LC/MS data.

Strengths:

The main strength is that GromovMatcher achieves significant performance metrics compared to other existing methods. The authors have done extensive comparisons to claim that GromovMatcher performs well.

Weaknesses:

There are two weaknesses.

(1) When the number of features is reduced the precision drops to ~0.8.

We would like to clarify that this drop in precision occurs in the challenging setting where only a small proportion of metabolites are shared between both datasets (e.g., the overlap – or proportion of shared features - was 25% in our simulation study). When two untargeted metabolic datasets share only 25% of their features, this is a challenging setting for any automated matching method as the vast majority 75% of the features in both datasets must remain unmatched.

In such settings, the reviewer correctly observes that the precision of GromovMatcher algorithms (GM and GMT) drops within the range of 0.80 - 0.85 (Figure 3b, top left panel). Such a precision of 0.8 or larger is still competitive compared with the alternative methods MetabCombiner (mC) and M2S whose precisions drop below 0.8 (see main text Fig. 3b, top left panel).

Precision is measured as the number of metabolite pairs correctly matched divided by all matches identified by a method. In other words, even in the challenging setting when the number of shared features (true matches) between both datasets is small (e.g. low 25% overlap), upwards of 80% of the feature matches found by GromovMatcher are correct which is a very encouraging result.

(2) How applicable is the method for other non-human datasets?

We thank the reviewer for raising this question. The crux of the matter concerning the application to animal data revolves around the hypothesis that correlations between metabolites in two different studies are preserved. Theoretically, the metabolome operates under similar principles in humans, governed by an underlying network of biochemical reactions. Consequently, in comparable human populations, the GM hypothesis is likely to hold to some extent.

However, in practice, application to animal data is more complicated. Animal studies tend to have smaller sample sizes and often stem from intervention-driven scenarios, such as mice subjected to specific diets or chemicals. This results in deliberate alterations in metabolic structures which makes finding two comparable animal studies less likely. To investigate the reviewer’s question, we have searched through the two predominant LC-MS dataset repositories (MetaboLights and NIH Metabolomics Workbench) but did not find any pairs of comparable animal studies due to the reasons mentioned above. One potential strategy to navigate this issue could entail regressing the metabolic intensities against the variables that notably differ between the two animal populations and running GM using the residual intensities. This would be an interesting direction for future research and additional validation would be needed to test the robustness of GM in this setting.

Reviewer #2 (Public Review):

Summary:

The goal of untargeted metabolomics is to identify differences between metabolomes of different biological samples. Untargeted metabolomics identifies features with specific mass-to-charge ratio (m/z) and retention time (RT). Matching those to specific metabolites based on the model compounds from databases is laborious and not always possible, which is why methods for comparing samples on the level of unmatched features are crucial.

The main purpose of the GromovMatcher method presented here is to merge and compare untargeted metabolomes from different experiments. These larger datasets could then be used to advance biological analyses, for example, for the identification of metabolic disease markers. The main problem that complicates merging different experiments is m/z and RT vary slightly for the same feature (metabolite).

The main idea behind the GromovMatcher is built on the assumption that if two features match between two datasets (that feature I from dataset 1 matches feature j from dataset 2, and feature k from dataset 1 matches feature l from dataset 2), then the correlations or distances between the two features within each of the datasets (i and k, and j and l) will be similar. The authors then use the Gromov-Wasserstein method to find the best matches matrix from these data.

The variation in m/z between the same features in different experiments is a user-defined value and it is initially set to 0.01 ppm. There is no clear limit for RT deviations, so the method estimates a non-linear deviation (drift) of RT between two studies. GromovMatcher estimates the drift between the two studies and then discards the matching pairs where the drift would deviate significantly from the estimate. It learns the drift from a weighted spline regression.

The authors validate the’performance of their GromovMatcher method by a validation experiment using a dataset of cord blood. They use 20 different splits and compare the GromovMatcher (both its GM and GMT iterations, whereby the GMT version uses the deviation from estimated RT drift to filter the matching matrix) with two other matching methods: M2S and metabCombiner.

The second validation was done using a (scaled and centered) dataset of metabolics from cancer datasets from the EPIC cohort that was manually matched by an expert. This dataset was also used to show that using automatic methods can identify more features that are associated with a particular group of samples than what was found by manual matching. Specifically, the authors identify additional features connected to alcohol consumption.

Strengths:

I see the main strength of this work in its combination of all levels of information (m/z, RT, and higher-order information on correlations between features) and using each of the types of information in a way that is appropriate for the measure. The most innovative aspect is using the Gromov-Wasserstein method to match the features based on distance matrices.

We thank the reviewer for acknowledging this strength of our proposed GromovMatcher method.

The authors of the paper identify two main shortcomings with previously established methods that attempt to match features from different experiments: a) all other methods require fine-tuning of user-defined parameters, and, more importantly, b) do not consider correlations between features. The main strength of the GromovMatcher is that it incorporates the information on distances between the features (in addition to also using m/z and RT).

Weaknesses:

The first, minor, weakness I could identify is that there seem not to be plenty of manually curated datasets that could be used for validation.

We thank the reviewer for raising this issue concerning manually curated validation data.

Manually curated datasets available for validation purposes are indeed scarce. This stems from the laborious nature of matching features across diverse studies, hence the need for automatic matching methods. Our future strategy involves further validation of the GromovMatcher approach as more data becomes accessible in EPIC and other cohorts.

The scarcity of real-life publicly available datasets that can be used for validation purpose is the reason why we conducted an extensive simulation study (main text “Validation on ground-truth data” and Appendix 3). It is notably thorough, arguably more comprehensive than previous validations, utilizes real-life untargeted data, and imitates situations where data originates from distinct untargeted metabolomics studies, complete with realistic noise parameters encompassing RT, mz, and feature intensities. Our validation study comprehensively explores the performance of GromovMatcher, M2S, and metabCombiner, including in challenging realistic settings where there is a nonlinear drift in retention times, varying levels of feature overlaps between studies, normalizations of feature intensities, as well as imbalances in the number of features and samples present in the studies being matched.

The second is also emphasized by the authors in the discussion. Namely, the method as it is set up now can be directly used only to compare two datasets.

This is indeed a limitation that is common to all three methods considered in this paper. However, all these methods, GromovMatcher, M2S, and metabCombiner, can still be used to compare and pool multiple datasets using a multi-step procedure. Namely, this can be done by designating a 'reference' dataset and aligning all studies to it one by one. We take this exact approach in our paper when aligning the CS, HCC, and PC studies of the EPIC data in positive mode (main text “Application to EPIC data”). Namely, the HCC and PC studies are both aligned to the CS study by running GromovMatcher twice, and after obtaining these matchings, our analysis is restricted to those features in HCC and PC that are present in the CS study.

After the reviewer’s comment, we have added an additional sensitivity analysis in Appendix 5, to compare the results produced by GromovMatcher depending on the choice of the reference study. Namely, setting the reference study to either the CS study or the HCC study, GromovMatcher identified 706 and 708 common features respectively, with an overlap of 640 features. This highlights that the choice of reference does matter to some extent. In our original analysis of the EPIC data, choosing CS as the reference was motivated by the fact that CS had the largest sample size (compared to HCC and PC) and a subset of features in HCC and PC were already matched by experts to the CS study which we could use for validation (see Loftfield et al. (2021). J Natl Cancer Inst.).

As mentioned in the discussion section of our manuscript, the recently proposed multimarginal Gromov-Wasserstein algorithm (Beier, F., Beinert, R., & Steidl, G. (2023). Information and Inference) could potentially allow multiple metabolomic studies to be matched using one optimization routine (e.g. without the designation of a ‘reference study’ for matching). We have not explored this possibility in depth yet as fast numerical methods for multimarginal GW are still in their infancy. Also, such multimarginal methods rely on the computation and storage of coupling or matching matrices that are tensors where the number of dimensions is equal to the number of datasets being matched. Therefore, multimarginal methods have large memory costs, which currently precludes their application for the matching of multiple metabolomics datasets.

Reviewer #2 (Recommendations For The Authors):

(1) I was struggling with the representation used in Figure 3a. The gray points overlayed over the green points on a straight line are difficult to visually quantify. I found that my eyes mainly focused on the pattern of the red dots.

Figure 3a has been modified to improve visual clarity. Namely we have consistently reordered the rows and columns of the coupling matrices such that the true positive matches (green points) are spatially separated from the false negative matches (red points). Now the fraction of true positive and false negative matches can be appreciated much more clearly by eye in Figure 3a.

(2) I would also like to add the caveat that I cannot judge whether the authors used the other two methods that they compare with GromovMatcher (the M2S and metabCombiner) optimally. But I also do not see any evidence that they did not. Hopefully one of the other reviewers can address that.

We appreciate the reviewer for highlighting the comparison of our approach GromovMatcher to the other existing methods M2S and MetabCombiner (mC). Both M2S and mC depend on tens of hyperparameters each with a discrete or continuous set of values that must be properly optimized to infer accurate matchings between dataset features. We detail in Appendix 2 how the hyperparameters of the M2S and mC methods are optimally tuned to achieve the best possible performance on the validation ground-truth data. Namely, both in the simulation study and on EPIC data, we grid-search over all important hyperparameters in the M2S and mC methods and choose those parameter combinations that result in the highest F1 score, averaged over 20 random trials. We remark that no such hyperparameter optimization was performed for our GromovMatcher method. As shown in Figures 3 and 4 of the main text, we find that GromovMatcher outperforms M2S and mC even in these cases when the hyperparameters of M2S and mC are tuned to predict optimal feature matchings.

Given the large combinatorial space of hyperparameter choices, we believe we have thoroughly tested the important hyperparameter combinations that users of M2S and mC would be likely to explore in their own research.

(3) Validation

(3a) The first validation is done on a split cord blood dataset. I could not clearly see from the paper how sensitive the result is to the dataset split.

We are grateful for the reviewer’s question and have included new experiments in Appendix 3 which show how the results of GromovMatcher, M2S, and MetabCombiner are affected by the dataset split. In our original manuscript, our validation ground-truth experiment began with an untargeted metabolomic dataset consisting of n = 499 samples and p = 4,712 metabolic features which is split equally into two datasets consisting of an equal number of samples n1 = n2 and an equal number of metabolic features p1 = p2. The features of these equal-sized datasets would then be matched by our method.

Now in Appendix 3 (Figs. 1-3) we show the sensitivity of all three alignment methods (GromovMatcher, M2S, and MetabCombiner) when we vary the fraction of samples in dataset 1 over dataset 2 given by n1/ n2, the overlap in shared features between both datasets, and the fraction of metabolic features in dataset 1 that are not present in dataset 2 which affects the feature sizes of both datasets p1/ p2. We find that all alignment methods are able to maintain a consistent precision and recall score when these three dataset split parameters are varied. GromovMatcher achieves a higher precision and recall than M2S and MetabCombiner for all choices of dataset split, agreeing with the validation experiment results from the main text (see main text Fig. 3). All three methods tested decrease in precision (without dropping in recall) when dataset 1 and dataset 2 contain an equal number of unshared features (e.g. when p1 = p2). Therefore, these sensitivity experiments in Appendix 3 show that our results in the main text are performed in the most challenging setting for the dataset split.

(3b) The second validation was done using a (scaled and centered) dataset of metabolics from cancer datasets from the EPIC cohort that was manually matched by an expert. Here the authors observe that metabCombiner has good precision, but lags in recall. And M2S has a very similar performance to GromovMatcher. The authors explain this by the fact that the drift in RT between the two experiments is mostly linear and thus does not affect the M2S performance. Can the authors find a different validation dataset where the drift in RT is not linear? If yes, it would be interesting to add it to the paper.

We thank the reviewer for raising this question. As mentioned above, curated validation datasets such as the EPIC study analyzed in our paper are very rare and we do not currently have a validation study with a nonlinear retention time drift.

Nevertheless, we performed an additional analysis of simulated data (reported in Appendix 2 – “M2S hyperparameter experiments” and Appendix 2 – Table 1) that demonstrates the decrease in M2S performance when the simulated drift is nonlinear. As presented in Appendix 2 – Table 1, in a low overlap setting with a linear drift which corresponds to the EPIC data, precision and recall were 0.831 and 0.934 respectively, instead of 0.769 and 0.905 in the main analysis where the drift was nonlinear.

Version 2

Abstract

Abstract

Untargeted metabolomic profiling through liquid chromatography-mass spectrometry (LC-MS) measures a vast array of metabolites within biospecimens, advancing drug development, disease diagnosis, and risk prediction. However, the low throughput of LC-MS poses a major challenge for biomarker discovery, annotation, and experimental comparison, necessitating the merging of multiple datasets. Current data pooling methods encounter practical limitations due to their vulnerability to data variations and hyperparameter dependence. Here we introduce GromovMatcher, a flexible and user-friendly algorithm that automatically combines LC-MS datasets using optimal transport. By capitalizing on feature intensity correlation structures, GromovMatcher delivers superior alignment accuracy and robustness compared to existing approaches. This algorithm scales to thousands of features requiring minimal hyperparameter tuning. Applying our method to experimental patient studies of liver and pancreatic cancer, we discover shared metabolic features related to patient alcohol intake, demonstrating how GromovMatcher facilitates the search for biomarkers associated with lifestyle risk factors linked to several cancer types.

Introduction

Untargeted metabolomics is a powerful analytical technique used to identify and measure a large number of metabolites in a biological sample without preselecting targets Patti (2011). This approach allows for a comprehensive overview of an individual’s metabolic profile, provides insights into the biochemical processes involved in cellular and organismal physiology Wishart (2019); Pirhaji et al. (2016), and allows for the exploration of how environmental factors impact metabolism Rappaport et al. (2014); Bedia (2022). It creates new opportunities to investigate health-related conditions, including diabetes Wang et al. (2011), inflammatory bowel diseases Franzosa et al. (2019), and various cancer types Loftfield et al. (2021); Li et al. (2020). However, a major challenge in biomarker discovery, metabolic signature identification and other untargeted metabolomic analyses lies in the low throughput of experimental data, necessitating the development of efficient pooling algorithms capable of merging datasets from multiple sources Loftfield et al. (2021).

A common experimental technique in untargeted metabolomics is liquid chromatography-mass spectrometry (LC-MS) which assembles a list of thousands of unlabeled metabolic features characterized by their mass-to-charge ratio (m/z), retention time (RT) Zhou et al. (2012), and intensity across all biological samples. Combining LC-MS datasets from multiple experimental studies remains challenging due to variation in the m/z and RT of a feature from one study to another Zhou et al. (2012); Ivanisevic and Want (2019). This problem is further compounded by differing instruments and analytical protocols across laboratories, resulting in seemingly incompatible metabolomic datasets.

Manual matching of metabolic features can be a laborious and error-prone task Loftfield et al. (2021). To address this challenge, several automated methods have been developed for metabolic feature alignment. One such method is MetaXCMS, which matches LC-MS features based on user-defined m/z and RT thresholds Tautenhahn et al. (2011). More advanced tools use information on feature intensities measured in samples. For instance, PAIRUP-MS uses known shared metabolic features to impute the intensities of all features from one dataset to another Hsu et al. (2019). MetabCombiner Habra et al. (2021) and M2S Climaco Pinto et al. (2022) compare average feature intensities, along with their m/z and RT values, to align datasets without requiring extensive knowledge of shared features. These automated alignment methods have accelerated our ability to pool and annotate datasets as well as extract biologically meaningful biomarkers. However, they demand substantial fine-tuning of user-defined parameters and ignore correlations among metabolic features which provide a wealth of additional information on shared features.

Here we introduce GromovMatcher, a user-friendly flexible algorithm which automates the matching of metabolic features across experiments. The main technical innovation of GromovMatcher lies in its ability to incorporate the correlation information between metabolic feature intensities, building upon the powerful mathematical framework of computational optimal transport (OT) Peyré et al. (2019); Villani (2021). OT has proven effective in solving various matching problems and has found applications in multiomics analysis Demetci et al. (2022), cell development Schiebinger et al. (2019); Yang et al. (2020), and chromatogram alignment Skoraczyński et al. (2022). Here we leverage the Gromov-Wasserstein (GW) method Mémoli (2011); Solomon et al. (2016), which matches datasets based on their distance structure and has been seminally applied to spatial reconstruction problems in genomics Nitzan et al. (2019). GromovMatcher builds upon the GW algorithm to automatically uncover the shared correlation structure among metabolic feature intensities while also incorporating m/z and RT information in the final matching process.

To assess the performance of GromovMatcher, we systematically benchmark it on synthetic data with varying levels of noise, feature overlap, and data normalizations, outperforming prior state-of-the-art methods of metabCombiner Habra et al. (2021) and M2S Climaco Pinto et al. (2022). Next we apply GromovMatcher to align experimental patient studies of liver and pancreatic cancer to a reference dataset and associate the shared metabolic features to each patient’s alcohol intake. Through these efforts, we demonstrate how GromovMatcher data pooling improves our ability to discover biomarkers of lifestyle risk factors associated with several types of cancer.

Results

GromovMatcher algorithm

GromovMatcher uses the mathematical framework of OT to find all matching metabolic features between two untargeted metabolomic datasets (Fig. 1). It accepts two LC-MS datasets with possibly different numbers of metabolic features and samples. Each feature, fxi. in Dataset 1 and fyj in Dataset 2, is identified by its m/z, RT, and vector of feature intensities across samples (Fig. 1a). The primary tenet of GromovMatcher is that shared metabolic features have similar correlation patterns in both datasets and can be matched based on the distance/correlations between their feature intensity vectors. Specifically, GromovMatcher computes the pairwise distances between the feature intensity vectors of each metabolic feature in a dataset and saves them into a distance matrix, one per dataset (Fig. 1b). In practice, we use either the Euclidean distance or the cosine distance (negative of correlation) to perform this step (Methods). The resulting distance matrices contain information about the feature intensity similarity within each study. Using optimal transport, we can deduce shared subsets of metabolic features in both datasets which have corresponding feature intensity distance structures.

An optimal transport approach for combining untargeted metabolomics datasets (GromovMatcher).

(a) Inputs are two LC-MS datasets of unlabeled metabolic features (rows) identified by their m/z, RT, and feature intensities across biospecimen samples. Both studies can have differing numbers of metabolic features and samples. (b) In both datasets, the intensities across samples of each metabolic feature are formed into a vector and Euclidean distances between these feature vectors are computed and stored in a distance matrix. (c) Based on the technique of optimal transport, the unbalanced GW algorithm learns a coupling matrix Π ˜ that places large weights Π ˜ ij ≥0 when fxi and fyy likely correspond to the same metabolic feature. It optimizes Π ˜ to match features with similar pairwise distances (red outlined boxes) whose m/z ratios are close. (d) The final step of GromovMatcher plots the retention times of features from both datasets against each other and fits a spline interpolation f weighted by the estimated coupling weights n. This retention time drift function is then used to set all entries Π ˜ ij to zero for those outlier pairs (fxi, fyy) which exceed twice the median absolute deviation (MAD) around Π ˜ (green highlighted region). Finally, the coupling matrix Π ˜ is filtered and/or thresholded to obtain a refined coupling Π ˜ which is then binarized to obtain a one-to-one matching M between a subset of metabolite pairs in both datasets.

OT was originally developed to optimize the transportation of soil for the construction of forts Monge (1781) and was later generalized through the language of probability theory and linear programming Kantorovich (2006), leading to efficient numerical algorithms and direct applications to planning problems in economics. The ability of OT to efficiently match source to target locations found applications in data science for the alignment of distributions Courty et al. (2017); Alvarez-Melis et al. (2019) and was generalized by the Gromov-Wasserstein (GW) method Peyré et al. (2016); Alvarez-Melis and Jaakkola (2018) to align datasets with features of differing dimensions.

In practice, a sizeable fraction of the metabolic features measured in one study may not be present in the other. Hence, in most cases only a subset of features in both datasets can be matched. Recent GW formulations for unbalanced matching problems Sejourne et al. (2021) allow for matching only subsets of metabolic features with similar intensity structures (Fig. 1c). To incorporate additional feature information, we modify the optimization objective of unbalanced GW to penalize feature matches whose m/z differences exceed a fixed threshold (Methods, Appendix 1). The optimization of this objective computes a coupling matrix Π ˜ where each entry Π ˜ ij ≥0 indicates the level of confidence in matching metabolic feature fxi in Dataset 1 to fyj in Dataset 2.

Differences in experimental conditions can induce variations in RT between datasets that can be nonlinear and large in magnitude Zhou et al. (2012); Climaco Pinto et al. (2022); Habra et al. (2021). In the spirit of previous methods for LC-MS batch or dataset alignment Smith et al. (2006); Brunius et al. (2016); Liu et al. (2020); Vaughan et al. (2012); Habra et al. (2021); Climaco Pinto et al. (2022); Skoraczynski et al. (2022), the learned coupling Π ˜ is used to estimate a nonlinear map (drift function) between RTs of both datasets by weighted spline regression, which allows us to filter unlikely matches from the coupling matrix to obtain a refined coupling matrix Π ˜ (Fig. 1d, Methods). An optional thresholding step removes matches with small weights from the coupling matrix. The final output of GromovMatcher is a binary matching matrix M where Mij is equal to 1 if features fxi and fyj are matched and 0 otherwise. Throughout the paper, we refer to the two variants of GromovMatcher, with and without the optional thresholding step as GMT and GM respectively.

Validation on ground-truth data

We first evaluate the performance of GromovMatcher using a real-world untargeted metabolomics study of cord blood across 499 newborns containing 4,712 metabolic features characterized by their m/z, RT, and feature intensities Alfano et al. (2020). To generate ground-truth data, we randomly divide the initial dataset into two smaller datasets sharing a subset of features (Fig. 2). We simulate diverse acquisition conditions by adding noise to the m/z and RT of dataset 2, and to the feature intensities in both datasets. Moreover, we introduce an RT drift in dataset 2 to replicate the retention time variations observed in real LC-MS experiments (Methods and Materials). For comparison, we also test M2S Climaco Pinto et al. (2022) and metabCombiner Habra et al. (2021), both of which use m/z, RT, and median or mean feature intensities to match features (Fig. 3). MetabCombiner is supplied with 100 known shared metabolic features to automatically set its hyperparameters, while M2S parameters are manually fine-tuned to optimize the F1-score in each scenario (Appendix 2). We assess the performance of GM, GMT, metabCombiner, and M2S across 20 randomly generated dataset pairs in terms of their precision (fraction of true matches among the detected matches) and recall/sensitivity (fraction of true matches detected) averaged across 20 dataset pairs.

To investigate how the number of shared features affects dataset alignment, we generate pairs of LC-MS datasets with low, medium, and high feature overlap (25%, 50%, and 75%), while maintaining a medium noise level (Methods). Here we find that GM and GMT generally outperform existing alignment methods, with a recall above 0.95 while metabCombiner and M2S tend to be less sensitive (Fig. 3b). All methods drop in precision as the feature overlap is decreased, with GM and GMT still maintaining an average precision above 0.8.

Simulated data for testing untargeted metabolomics alignment methods.

(a) Initial LC-MS dataset taken from the EXPOsOMICS project with m/z, RT, and feature intensities of p = 4, 712 metabolites identified in cord blood across n = 499 newborns. (b) Newborns (rows) are split into two disjoint groups of sizes n1 = 249 and n2 = 250 respectively and metabolic features (columns) are split into two equal groups of size p1 = p2 with overlap λp where λ = 0.25, 0.5, 0.75 (Methods). Datasets are perturbed by additive noise of magnitude (σM, σRT, σFl) and a nonlinear drift f (x) is applied to the RTs of dataset 2. (c) The two resulting datasets share λ = 25%, 50%, or 75% of the original dataset’s metabolic features.

Next we evaluate all four methods at low, moderate, and high noise levels for pairs of datasets with 50% overlap in their features (Methods). Our results show that GMT, GM, and M2S maintain an average recall above 0.89, while metabCombiner’s recall drops below 0.6 for high noise. At large noise levels, RT drift estimation becomes more challenging, leading to a higher rate of false matches between metabolites (lower precision) for all four methods (Fig. 3 - figure supplement 1). Nevertheless, GMT obtains a high average precision and recall of 0.86 and 0.92 respectively.

A notable difference between GM, metabCombiner, and M2S lies in their use of feature intensities. MetabCombiner expects that the mean feature intensity rankings are identical across studies, while M2S assumes that shared features have similar median intensities. In contrast, GM uses both the mean feature intensities and their variances and covariances. In practice, differences in experimental assays or study populations can lead to greater variation in feature intensities, making matchings based on these statistics less reliable. Centering and scaling the feature intensities to unit variance avoids potential biases arising from inconsistent feature intensity magnitudes, but preserves correlations that GM leverages.

Exploring this further, we test how sensitive all four methods are to centering and scaling of feature intensities. MetabCombiner and M2S are tuned using the same methodology as for non-centered and non-scaled data. For M2S, we match features solely based on their m/z and RT. In this experiment (Figure 3 - figure supplement 2), the absence of intensity magnitude information significantly affects metabCombiner’s performance and, to a lesser extent, M2S. GM and GMT still obtain accurate matchings, due to their use of correlation structures which are preserved under centering and scaling.

Application to EPIC data

Next, we apply GM, metabCombiner and M2S to align datasets from the European Prospective Investigation into Cancer and Nutrition (EPIC) cohort, a prospective study conducted across 23 European centers. EPIC comprises more than 500,000 participants who provided blood samples at recruitment Riboli et al. (2002). Untargeted metabolomics data were successively acquired in several studies nested within the full cohort.

In the present work, we use LC-MS data from the EPIC cross-sectional (CS) study Slimani et al. (2003) and two matched case-control studies nested within EPIC, on hepatocellular carcinoma (HCC) Stepien et al. (2016, 2021) and pancreatic cancer (PC) Gasull et al. (2019). LC-MS untargeted metabolomic data were acquired at the International Agency for Research on Cancer, making use of the same platform and methodology (Methods). The number of samples and features in each study is displayed in Fig. 4a.

Comparison of MetabCombiner, M2S, and GromovMatcher on simulated data.

(a) Ground-truth matchings, and matchings inferred by metabCombiner, M2S, GM, and GMT. Pairs of datasets are generated for three levels of overlap (low, medium and high), with a medium noise level (Methods). Matches correctly recovered (true positives) are represented in green. True matches that are not recovered (false negatives) are highlighted in grey. Incorrect matches (false positives) are plotted in red. Features in rows and columns of matching matrices are reordered for visual clarity. (b) Average precision and recall on 20 randomly generated pairs of datasets, for three levels of overlap (low, medium and high) with a medium noise level.

Loftfield et al. Loftfield et al. (2021) previously matched features from the CS, HCC, and PC studies in EPIC for alcohol biomarker discovery. The authors first identified 205 features (163 in positive and 42 in negative mode) associated with alcohol intake in the CS study. These features were then manually matched by an expert to features in both the HCC and PC studies (Methods). In our analysis, we use these features as a validation set and compare each method’s matchings to the expert manual matchings on this subset. Due to the imbalance between the number of positive and negative mode features in the validation subset, our main analysis focuses on the alignment results of CS with HCC and CS with PC in positive mode. We delegate the matching results between the negative mode studies to Appendix 4.

In this section, we use the same settings for GM as in our simulation study, and do not apply an additional thresholding step. The parameters of metabCombiner and M2S are calibrated using the validation subset as prior knowledge (Appendix 2).

Preliminary analysis of the validation subset reveals inconsistencies in the mean feature intensities (Figure 4 - figure supplement 1), but Figure 4b shows that on centered and scaled data, the 90 expert matched features shared between the CS and HCC studies have similar correlation structures. Hence, to avoid potential errors we center and scale the feature intensities which improves the performance of all three methods tested below (Appendix 4, Appendix 4 - Table 1).

Hepatocellular carcinoma

Here we analyze the quality of the matchings obtained by GM, M2S, and metabCombiner between the CS and HCC datasets in positive mode. Both GM and M2S identify approximately 1000 shared features while metabCombiner finds a smaller number of about 700 shared features. We refer the reader to Figure 4- figure supplement 2a for the precise matched feature sizes and details on the agreement between the feature matchings of all three methods.

We evaluate the performance of metabCombiner, M2S, and GM on the validation subset in positive mode (Fig. 4c), which consist of 90 features from the CS study manually matched to features from the HCC study and 73 features specific to the CS study. MetabCombiner demonstrates precise matching but lacks sensitivity. M2S’s precision and recall are comparable with GM, in contrast to its performance on simulated data. This can be attributed to the RT drift shape between the CS and HCC studies (Appendix 2), which is estimated to be close to linear (Figure 4 - figure supplement 3). Because the parameters of M2S are fine-tuned in the validation subset, it is able to learn this linear drift and apply tight RT thresholds to achieve accurate matchings. In contrast to metabCombiner and M2S, the GM algorithm is not given any prior knowledge of the validation subset, and nevertheless demonstrates the highest precision and recall rates of the three methods (Fig. 4c). Figure 4b shows how GM recovers the majority of the expert matched pairs by leveraging the shared correlations.

Pancreatic cancer

Matching features between the CS and PC studies in positive mode, GM and M2S identify approximately 1000 common features, while metabCombiner detects approximately 600 matches (Figure 4- figure supplement 2b). We examine the performance of all three methods on the validation subset consisting of 66 manually matched features between CS and PC along with 97 features specific to the CS study. As before, GM and M2S have high recall while the recall of metabCombiner is less than 0.5.

A decrease in precision is observed for both GM and M2S compared to the previous CS-HCC matchings. We therefore manually inspect the false positive matches; the set of CS features matched by the method to the PC study but explicitly examined and left unmatched in the expert manual matching. Assessing the GM results, we identify 7 false positive feature matches. Upon secondary inspection, 3 pairs are revealed as correct matches that were not initially identified in the expert matching. M2S finds 11 false positive matches which include the 7 false positives recovered by GM. Manual examination of the 4 remaining pairs reveals 2 clear mismatches. These results highlight the advantage of using automated methods for data alignment, as both GM and M2S detect correct matches that were not identified by experts, with GM being more precise than M2S.

Application of GromovMatcher and comparison to existing methods on EPIC dataset.

(a) Dimensions of the three EPIC studies used. For each ionization mode, the cross-sectional (CS) study is aligned successively with the hepatocellular carcinoma (HCC) study and the pancreatic cancer (PC) study. (b) Demonstration of expert manual matching and GromovMatcher (GM) matching between the CS and HCC studies in positive mode. Experts manually match 90 features from Loftfield et al. (2021) and the correlation matrices of these features in both datasets have similar structure (bottom two matrices). GM discovers 996 shared features between the CS and HCC datasets which have similar correlation structure (top two matrices). We validate that 88 of the 90 features from the manually expert matched subset are contained in the set of features matched by GM. (c) Performance of metabCombiner (mC), M2S and GM in positive mode. Precisions and recalls are measured on a validation subset of 163 manually examined features, and 95% confidence intervals are computed using modified Wilson score intervals. (d) Performance of mC, M2S and GM in negative mode. Precision and recall are measured on a validation subset of42 manually examined features, and 95% confidence intervals are computed using modified Wilson score intervals. See Table 2 and Table 3 for exact precisions, recalls, and confidence intervals in positive and negative mode respectively.

Illustration for alcohol biomarker discovery

Loftfield et al. Loftfield et al. (2021) identified biomarkers of habitual alcohol intake by first performing a discovery step, where they examined the relationship between alcohol intake and metabolic features in the CS study. They then manually matched the significant features in CS to features from the HCC and PC studies, and repeated the analysis with samples from the HCC and PC studies to determine whether the association with alcohol intake persisted. This led to the identification of 10 features possibly associated with alcohol intake (Fig. 5a).

To extend this analysis and illustrate the benefit of GM automatic matching for biomarker discovery, we use GM to pool features from the CS, HCC, and PC studies, and examine the relationship between metabolic features and alcohol intake in the pooled study (Methods and Fig. 5b).

Applying an FDR correction on the pooled study, we identify 243 features associated with alcohol intake, including 185 features consistent with the discovery step of Loftfield et al. Loftfield et al. (2021), and 55 newly discovered features (Fig. 5c). Using the more stringent Bonferroni correction on the pooled data, we identify 36 features shared by all three studies that are significantly associated with alcohol intake. These features include all 10 features identified in Loftfield et al. (Fig. 5c). These findings highlight the potential benefits of using GM automatic matching for biomarker discovery in untargeted metabolomics data. Additional information regarding the methodology and findings of our GM and Loftfield et al. analyses can be found in Methods and Appendix 4.

Discussion

LC-MS metabolomics has emerged as an increasingly powerful tool for biological and biomedical research, offering promising opportunities for epidemiological and clinical investigations. However, integrating data from different sources remains challenging. To address this issue, we introduce GromovMatcher, a method based on optimal transport that automatically aligns LC-MS data from pairs of studies. Our method exhibits superior performance on both simulated and real data when compared to existing approaches. Additionally, it presents a user-friendly interface with few hyperparameters.

While GromovMatcher is robust to noise and variations in data, it may face limitations when aligning LC-MS studies from populations with different characteristics, where the correlation structures between features may be inconsistent across studies. In this case, the base assumption of GromovMatcher can be relaxed by focusing on subsamples with similar characteristics, as exemplified in a recent study Gomari et al. (2022).

A current limitation is that GromovMatcher does not account for more than two datasets simultaneously, although this can be overcome by aligning multiple studies to a chosen reference dataset, as demonstrated in our biomarker experiments. The extension of Gromov-Wasserstein to multiple distributions Beier et al. (2022) is another promising approach for generalizing GromovMatcher to multiple dataset alignment. Further improvements can be made by incorporating existing knowledge about the studies being matched, such as known shared features, samples in common, or MS/MS data.

The results obtained from GromovMatcher are highly promising, opening the door for various analyses of metabolomic datasets acquired in different experimental laboratories. Here we demonstrated the potential of GromovMatcher in expediting the combination and meta-analysis of data for biomarker and metabolic signature discovery. The matchings learned by GromovMatcher also allow for comparison between experimental protocols by assessing the drift in m/z, RT, and feature intensities across studies. Finally, inter-institutional annotation efforts can directly benefit from incorporating this method to transfer annotations between aligned datasets. Bridging the gap between otherwise incompatible LC-MS data, GromovMatcher enables seamless comparison of untargeted metabolomics experiments.

Comparison of GromovMatcher and Loftfield et al. Loftfield et al. (2021) analysis for alcohol biomarker discovery on EPIC data.

(a) Loftfield study implemented a discovery step, examining the relationship between alcohol intake and metabolic features in the CS study. The significant features in CS were manually matched to features from the HCC and PC and the analysis was repeated using samples from the HCC and PC studies. After this step, 10 features associated with alcohol intake were identified. (b) GromovMatcher analysis begins by matching features from CS study to HCC and PC studies respectively (top blue, yellow, and red boxes). Samples corresponding to each CS feature are combined with the samples of its matched feature in the HCC study, PC study, or both. This generates a larger pooled data matrix with the same number of features as the CS study but with more samples pooled across the three original studies (center matrix). Because some features in the CS study may not have matches in HCC or PC, the corresponding entries in the pooled matrix are set to NaN/missing values (white regions in matrix). Each column/feature in this matrix is statistically tested for association with alcohol intake (ignoring missing values) and an FDR or a stricter Bonferroni correction is performed to retain only a subset of features from the pooled study that have a strong association. (c) Venn diagrams show intersection of feature sets (in positive and negative mode) found to be associated with alcohol intake by one of the four different analyses.

Methods and Materials

GromovMatcher method overview

GromovMatcher accepts as input two feature tables from separate LC-MS untargeted metabolomics studies. Each feature table for dataset 1 and dataset 2 consists of n1, n2 biospecimen samples respectively and p1, p2 metabolic features respectively detected in the study. Features in dataset 1 are given the label fxi for i = 1,… ,p1. Every feature is characterized by a mass-to-charge ratio (m/z) denoted by n i x , a retention time (RT) denoted by R T i x , and a vector of intensities across all samples written as X i ∈ ℝ n 1 . Similarly, features in dataset 2 are labeled as fyj for j = 1,… ,p2 and are characterized by their m/z m j y , retention time R T j y , and a vector of intensities across all samples Y i ∈ ℝ n 2 .

Our goal is to identify pairs of indexes (i, j) with i ϵ {1, . . . , p1} and j ϵ {1, . . ., p2}, such that fxi and fyj correspond to the same metabolic feature. More formally, we aim to identify a matching matrix M∈ {0,1} p 1 × p 2 such that Mij = 1 if fxi and fyj correspond to the same feature, hereafter referred to as matched features. Otherwise, we set Mij = 0.

Because the m/z and RT values of metabolomic features are often noisy and subject to experimental bias, our matching algorithm leverages metabolite feature intensities Xi,Yj to produce accurate dataset alignments. The GromovMatcher method is based on the idea that signal intensities of the same metabolites measured in two different studies should exhibit similar correlation structures, in addition to having compatible m/z and RT values. Here we define the Pearson correlation for vectors u, v, ϵ ℝn as

where we define

as the mean value, Euclidean norm and inner product respectively. If measurements Xi,Yj correspond to the same underlying feature, and similarly, measurements Xk,Yl, share the same an underlying feature, we expect that

This idea that the feature intensities of shared metabolites have the same correlation structure in both datasets also holds more generally for distances, under a suitable choice of distance. For example, the correlation coefficient corr( u, V) can be turned into a dissimilarity metric by defining

commonly referred to as the cosine distance. Preservation of feature intensity correlations then trivially amounts to the preservation of cosine distances.

Another classical notion of distance between vectors u, v, ϵ ℝn is the normalized Euclidean distance

which is equal to the cosine distance (up to constants) when the vectors u, v are centered and scaled to have zero mean and a standard deviation of one. The Euclidean distance depends on the magnitude or mean intensity of metabolic features, and hence is a useful metric for matching metabolites as long as these mean feature intensities are reliably collected.

To summarize, the main tenant of GromovMatcher is that if measurements Xl,Yj correspond to the same feature and Xk, Yl, correspond to the same feature, then for suitably chosen distances d x : ℝ n 1 × ℝ n 1 →ℝ and d y : ℝ n 2 × ℝ n 2 →ℝ , these distances are preserved

across both datasets. In this paper, the distances dx,dy are taken to be the normalized Euclidean distances in (5). We take care to specify those experiments where the metabolic features X and Y are centered and scaled. In these cases, implicitly the Euclidean distance between normalized feature vectors becomes the cosine distance (4) between the original (unnormalized) feature vectors.

Unbalanced Gromov–Wasserstein

The goal of GromovMatcher is to learn a matching matrix M∈{0,1}p1×p2 that gives an alignment between a subset of metabolites in both datasets. However, searching over the combinatorially large set of binary matrices would be an inefficient approach for dataset alignment. The mathematical framework of optimal transport Peyré et al. (2019) instead enlarges this space of binary matrices to the set of coupling matrices with real nonnegative entries ∈ ℝ + p 1 × p 2 . The entries ∏ij with large weights indicate that feature fxi in dataset 1 and feature fyj in dataset 2 are a likely match. Taking inspiration from (6), we minimize the following objective function

to estimate the coupling matrix ∏.

A standard approach is to optimize this objective over all coupling matrices ∏ under exact marginal constraints Π 1 =Π 1 p 2 , Π 2 = Π T 1 p 1 . Here we define 1n is the ones vector of length n, and X i i=1 p 1 , Y i i=1 p 2 denote the column and row sums of the coupling matrix. Objective (7) under these exact marginal constraints defines a distance between the two sets of metabolic feature vectors X i i=1 p 1 , Y i i=1 p 2 known as the Gromov–Wasserstein distance Mémoli (2011), a generalization of optimal transport to metric spaces. Note that for pairs Xi, Yj and Xk, Yj, for which dx(Xi, Xk) ≈ dy(Yj,Yl), the entries ∏lj, ∏kl, are penalized less and hence matches between features fxi, fyj and features fxk, fyl, are more favored. In our optimization, we avoid enforcing exact marginal constraints on the marginal distributions ∏1 p2 and ∏1p1 of our coupling matrix as this would enforce that all metabolites in both datasets are matched (Appendix 1). However, without any marginal constraints on the coupling ∏, the objective function (7) is trivially minimized by ∏ = 0, leaving all metabolites in both datasets unmatched.

To account for this, we follow the ideas of unbalanced Gromov–Wasserstein (UGW) Sejourne et al. (2021) and add three regularization terms to our objective

where ρ, ϵ > 0 and we define a = 1p1, b = 1p2. Here ⊗ denotes the Kronecker product. We define DKL as the Kullback-Leibler (KL) divergence between two discrete distributions μ,v∈ ℝ + p by

which measures the closeness of probability distributions.

The first two regularization terms in (8) enforce that the row sums and column sums of the coupling matrix ∏ do not deviate too much from a uniform distribution, leading our optimization to match as many metabolic features as possible. The magnitude of the regularizer ρ roughly enforces the fraction of metabolites in both datasets that are matched where large ρ implies most metabolites are matched across datasets. The final regularization term ϵ in (8) controls the smoothness (entropy) of the coupling matrix ∏ where larger values of ϵ encourage ∏ to put uniform weights on many of its entries, leading to less precision in the metabolite matches. However, increasing ϵ also leads to better numerical stability and a significant speedup of the alternating minimization algorithm used to optimize the objective function (Appendix 1). In our implementation, we set ρ and ϵ to the lowest possible values under which our optimization converges, with ρ = 0.05 and ϵ = 0.005.

Our full optimization problem can now be written as

The UGW objective function is optimized through alternating minimization based on the code of Sejourne et al. (2021) using the unbalanced Sinkhorn algorithm Séjourné et al. (2019) from optimal transport (Appendix 1).

Constraint on m/z ratios

Matched metabolic features must have compatible m / z so we enforce that ∏ ij =0 when m i x − m j y > m gap where m gap is a user-specified threshold. Based on prior literature Loftfield et al. (2021); Hsu et al. (2019); Climaco Pinto et al. (2022); Habra et al. (2021); Chen et al. (2021) we set mgap = 0.01ppm. Note that m gap is not explicitly used in (10) but is rather enforced in each iteration of our alternating minimization algorithm for the UGW objective (Appendix 1).

Unlike the m/z ratios discussed above, RTs often exhibit a non-linear deviation (drift) between studies so we cannot enforce compatibility of RTs directly in our optimization. Instead, in the following step of our pipeline we ensure matched metabolite pairs have compatible RTs by estimating the drift function and subsequently using it to filter out metabolite matches whose RT values are inconsistent with the estimated drift.

Estimation of the RT drift and filtering

Estimating the drift between RTs of two studies is a crucial step in assessing the validity of metabolite matches and discarding those pairs which are incompatible with the estimated drift.

Let Π ˜ ∈ ℝ + p 1 × p 2 be the minimizer of (10) obtained after optimization. We seek to estimate the RT drift function f: ℝ + → ℝ + which relates the retention times of matched features between the two studies. Namely, if feature fxi and feature fyj correspond to the same metabolic feature, then we must have that R T j y ≈f R T i x .

We propose to learn the drift f through the weighted spline regression

where Bn,k is the set of n-order B-splines with k knots. All pairs R T i x ,R T j y in objective (11) are weighted by the coefficients of Π ˜ so that larger weights are given to pairs identified with high confidence in the first step of our procedure. The order of the B-splines was set to n = 3 by default, while the number of knots k was selected by 10-fold cross-validation.

Pairs identified as incompatible with the estimated RT drift are then discarded from the coupling matrix. To do this, we first take the estimated RT drift f ^ , and the set of pairs S= i,j: Π ˜ i,j ≠0 recovered in Π ˜ . We then define the residual associated with (i, j) ϵ s as

The 95% prediction interval and the median absolute deviation (MAD) of these residuals are given by

where |S| is the size of S and the functions std and median denote the standard deviation and median respectively. Similar to the approach in Climaco Pinto et al. (2022), we create a new filtered coupling matrix Π ^ ∈ ℝ + p 1 × p 1 given by

where rthresh is a given filtering threshold. Following Habra et al. (2021), the estimation and outlier detection step can be repeated for multiple iterations, to remove pairs that deviate significantly from the estimated drift and improve the robustness of the drift estimation. In our main algorithm, we use two preliminary iterations where estimate the RT drift and discard outliers outside of the 95% prediction interval by setting rthresh = PI. We the re-estimate the drift and perform a final filtering step with the more stringent MAD by setting rthresh = 2 × MAD.

At this stage, it is possible for Π ˜ to still contain coefficients of very small magnitude. As an optional postprocessing step, we discard these coefficients by setting all entries smaller than τmax( Π ^ ) to zero, for some user-defined τ ϵ [0,1]. Lastly, a feature from either study could have multiple possible matches, since Π ˜ can have more than one non-zero coefficient per row or column. Although reporting multiple matches can be helpful in an exploratory context, for the sake of simplicity in our analysis, the final output of GromovMatcher returns a one-to-one matching, as we only keep those metabolite pairs (i,j) where the entry Π ^ ij is largest in its corresponding row and column. All nonzero entries of Π ˜ which do not satisfy this criterion are set to zero. Finally, we convert Π ˜ into a binary matching matrix M∈{0,1}p1×p2 with ones in place of its nonzero entries and this final output is returned to the user.

As a naming convention, we use the abbreviation GM for our GromovMatcher method, and use the abbreviation GMT when running GromovMatcher with the optional τ -thresholding step with τ = 0.3.

Metrics for dataset alignment

Every alignment method studied in this paper returns a binary partial matching matrix M∈{0,1}p1×p2 which has at most one nonzero entry in each row and column. Specifically, Mij = 1 if metabolic features i and j in both datasets correspond to each other and Mij = 0 otherwise. In our simulated experiments, we compare the partial matching M to a known ground-truth partial matching matrix M*∈{0,1}p1×p2 .

To do this, we first compute the number of true positives, false positives, true negatives, and false negatives as

where 1 denotes the indicator function. Then we use these values to compute the precision and recall as

Precision measures the fraction of correctly found matches out of all discovered metabolite matches, while recall, also know as sensitivity, measures the fraction of correctly matched pairs out of all truly matched pairs. These two statistics can be summarized into one metric called the F1-score by taking their harmonic mean

These three metrics, precision, recall, and the F1-score, are used throughout the paper to assess the performance of dataset alignment methods, both on simulated data where the ground-truth matching is known, and on the validation subset in EPIC, using results from the manual examination as the ground-truth benchmark.

Validation on simulated data

To assess the performance of GromovMatcher and compare it to existing dataset alignment methods, we simulate realistic pairs of untargeted metabolomics feature with known ground-truth matchings. This allows us to analyze the dependence of alignment methods on the number of shared metabolites, dataset noise level, and feature intensity centering and scaling.

Dataset generation

Our pairs of synthetic feature tables are generated from one real untargeted metabolomics study of 500 newborns within the EXPOsOMICS project, which uses reversed phase liquid chromatography-quadrupole time-of-flight mass spectrometry (UHPLC-QTOF-MS) system in positive ion mode Alfano et al. (2020). The original dataset is first preprocessed following the procedure detailed in Alfano et al. Alfano et al. (2020), resulting in p = 4,712 features measured in n = 499 samples available for subsequent analysis. Features and samples from the original study are then divided into two feature tables of respective size (n1 ,p1) and (n2,p2), with n1 + n2 = n and p1 ,p2 ≤ p. In order to do this, n1 = [n/2] randomly chosen samples from the original study are placed into dataset 1 and the remaining n2 = [n/2] samples from the original study are placed into dataset 2. Here [.] and [.] denote integer floor and ceiling functions. The features of the original study are randomly assigned to dataset 1, dataset 2, or both, allowing the resulting studies to have both common and study-specific features (Fig. 2). Specifically, for a fixed overlap parameter λ ϵ [0,1], we assign a random subset of ≈ λp features into both dataset 1 and dataset 2 while the remaining ≈ (1 – λp) features are divided equally between the two studies such that p1 = p2. We choose λ ϵ {0.25, 0.5, 0.75} corresponding to low, medium and high overlap. For more detailed information on how the dataset split is performed and for additional validation experiments with unbalanced dataset splits (e.g. n1 ≠ n2, p1 ≠ p2) we refer the reader to Appendix 3.

After generating a pair of studies, random noise is added to the m/z, RT and intensity levels of features in dataset 2 to mimic variations in data acquisition across two different experiments. The noise added to each m/z value in study 2 is sampled from a uniform distribution on the interval [–σM, σM] with σM = 0.01 Climaco Pinto et al. (2022). The RTs of dataset 2 are first deviated by the function f(x)=1.1x+1.3sin(1.2 x ) , corresponding to a systematic inter-dataset drift Habra et al. (2021); Climaco Pinto et al. (2022); Brunius et al. (2016). A uniformly distributed noise on the interval [–σRT, σRT] is added to the deviated RTs of dataset 2, with σRT ϵ {0.2, 0.5, 1} (in minutes) corresponding to low, moderate and high variations Climaco Pinto et al. (2022); Habra et al. (2021); Vaughan et al. (2012). Finally, we add a Gaussian noise N 0, σ FI 2 to the feature intensities of both studies where σFI is the scalar variance of the noise. This noise perturbs the correlation matrices of dataset 1 and dataset 2, making matching based on feature intensity correlations more challenging. We vary σFI over the set of values {0.1, 0.5, 1}.

Given this data generation process, we test the performance of the four alignment methods (M2S, metabCombiner, GM, and GMT) under the parameter settings described below.

Dependence on overlap

We first assess how the performance of the four methods is affected by the number of metabolic features shared in both datasets. For each value of λ = 0.25, 0.5, 0.75 (low, medium and high overlap), we randomly generate 20 pairs of datasets with noise on the m/z, RT and feature intensities set to σM = 0.01 , σRT = 0.5, σFI = 0.5. The precision and recall of each method at low, medium, and high overlap is recorded for each of the repetitions.

Noise robustness

Next we test the robustness to noise of each method by fixing the metabolite overlap fraction at λ = 0.5 and generating 20 random pairs of datasets at low (σRT = 0.2, σFI = 0.1), medium (σRT = 0.5, σFI = 0.5), and high σRT = 1, σFI = 1) noise levels. Similarly, the precision and recall of each method is saved for each noise level across the 20 repetitions.

Feature intensity centering and scaling

In order to test how all four methods are affected when the mean feature intensities and variance are not comparable across studies, we assess their performance when the feature intensities in both studies are mean centered and standardized to have unit standard deviation across all samples. We again generate 20 random pairs of datasets with medium overlap and medium noise, normalize the feature intensities in each pair of datasets, and compute the precision and recall of each method across the 20 repetitions.

EPIC data

We also evaluate our method on data collected within the European Prospective Investigation into Cancer and Nutrition (EPIC) cohort, an ongoing multicentric prospective study with over 500,000 participants recruited between 1992 and 2000 from 23 centers in 10 European countries, and who provided blood samples at the inclusion in the study Riboli et al. (2002). In EPIC, untargeted metabolomics data were successively acquired in several studies nested within the full cohort.

In the present work, we use untargeted metabolomics data acquired in three studies nested in EPIC, namely the EPIC cross-sectional (CS) study Slimani et al. (2003) and two matched case-control studies nested within EPIC, on hepatocellular carcinoma (HCC) Stepien et al. (2016, 2021) and pancreatic cancer (PC) Gasull et al. (2019), respectively. All data were acquired at the International Agency for Research on Cancer, making use of the same plateform and methodology: UHPLC-QTOF-MS (1290 Binary Liquid chromatography system, 6550 quadrupole time-of-flight mass spectrometer, Agilent Technologies, Santa Clara, CA) using reversed phase chromatography and electrospray ionization in both positive and negative ionization mode.

In a previous analysis aiming at identifying biomarkers of habitual alcohol intake in EPIC, the 205 features associated with alcohol intake in the CS study were manually matched to features in both the HCC and PC studies Loftfield et al. (2021). The results from this manual matching are presented in Table 1. This matching process was based on the proximity of m/z and RT, using a matching tolerance of ±15 ppm and ±0.2 min, and on the comparison of the chromatograms of features in a quality control samples from both studies.

Preprocessing

In the HCC and PC studies, samples corresponding to participants selected as cases in either study (i.e., participants selected in the study because of a diagnosis of incident HCC or PC) are excluded. Indeed, the metabolic profiles of participants selected as controls are expected to be more comparable across studies than those of cases, especially if certain features are associated with the risk of HCC or PC. Apart from this additional exclusion criterion, the untargeted metabolomics data of each study is pre-processed following the steps described in Loftfield et al. Loftfield et al. (2021), to eliminate unreliable features and samples, impute missing values and minimize technical variations in the feature intensity levels.

Alcohol biomarker discovery

Loftfield et al. Loftfield et al. (2021) used the untargeted metabolomics data of the CS, HCC and PC studies in their alcohol biomarker discovery study in EPIC, without being able to automatically match their common features and pool the 3 datasets. Instead, the authors first implemented a discovery step, examining the relationship between alcohol intake and metabolic features measured in the CS study and accounting for multiple testing using a false discovery rate (FDR) correction. This led to the identification of 205 features significantly associated with alcohol intake in the CS study. In order to gauge the robustness of these associations, the authors of Loftfield et al. Loftfield et al. (2021) then implemented a validation step using data from two independent test sets. The first test set was composed of data from the EPIC HCC and PC studies, while the second was derived from the Finnish Alpha-Tocopherol, Beta-Carotene Cancer Prevention (ATBC) study. The 205 features identified in the discovery step were manually investigated for matches in the EPIC test set, and 67 features were effectively matched to features in the HCC or PC study, or both. The authors then evaluated the association between alcohol intake and those 67 features, applying a more conservative Bonferroni correction to determine whether the association with alcohol intake persisted. This step led to the identification of 10 features associated with alcohol intake (Extended Data Fig. 5a). The second test set was then used to determine whether those 10 features were also significant in the ATBC population, which was indeed the case.

To conduct a more in-depth investigation of the matchings produced by the GromovMatcher algorithm, we build upon the analysis previously conducted by Loftfield et al. Loftfield et al. (2021) by exploring potential alcohol biomarkers using a pooled dataset created from the CS, HCC, and PC studies. Our goal is to assess whether pooling the data leads to increased statistical power and allows for the detection of more features associated with alcohol intake. Namely, we generate the pooled dataset by aligning a chosen reference dataset (CS study) with the HCC and PC studies successively using the GM matchings computed in both positive and negative mode (Methods and Extended Data Fig. 5b). Features that are not detected in either the HCC or PC studies are designated as ‘missing’ in the final pooled dataset for samples belonging to the respective studies where the feature is not found.

To evaluate the potential relationship between alcohol consumption and pooled metabolic features, we use a methodology akin to that of Loftfield et al. Loftfield et al. (2021). The self-reported alcohol intake data is adjusted for various demographic and lifestyle factors (age, sex, country, body-mass-index, smoking status and intensity, coffee consumption, and study) via the residual method in linear regression models. Feature intensities are also adjusted for technical variables (plate number and position within the plate) via linear mixed effect models. The significance of the association is assessed using correlation coefficients computed from the residuals for both selfreported alcohol intake and feature intensities. P-values are corrected using either false discovery rate (FDR) or Bonferroni correction to account for multiple testing. Corrected p-values less than 5% are considered significant.